---
title: "2.Fig1AEF_TileseqvsNGSvsDuplex"
author: "Haider Inam"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
# rm(list=ls())
library(stringr)
library(dplyr)
library(ggplot2)
library(plotly)
library(tictoc)
library(doParallel)
library(foreach)
library(RColorBrewer)
library("ggrepel")
library(reshape2)
library(ggpubr)
library(tidyr)
```

```{r cleanup}
source("code/resmuts_adder.R")
#Cleanup code for plotting
cleanup=theme_bw() +
  theme(plot.title = element_text(hjust=.5),
        panel.grid.major = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.background = element_blank(),
        # axis.line = element_line(color = "black"),
        axis.text = element_text(face="bold",color="black",size="11"),
        text=element_text(size=11,face="bold"),
        axis.title=element_text(face="bold",size="11"))
```
Doing an NGS, SSCS, DCS, and Tileseq comparison
The following data was taken from lane 18b sample 8 vs 10 duplex and tileseq data. This is region 1 (residues 242-322) of the ABL kinase sequenced with four different sequencing technologies. This sequencing data was generated using the scripts labeled ErrorRates_dataparser_region1.

## Plots for figures in manuscript


```{r,include=F}
#Cleanup code for plotting
source("code/compare_screens_archive.R")

source("code/plotting/cleanup.R")
source("code/cosmic_data_adder.R")
# source("code/compare_screens.R")

source("code/compare_screens_twosamples.R")
source("code/merge_samples_twosamples.R")
source("code/plotting/heatmap_plotting_function.R")
```


```{r}
# This sequencing data was generated using the scripts labeled ErrorRates_dataparser_region1.
data=read.csv("Figures/Figure_1/outputs/baf3_Imat_high_rep1vsrep2_duplex/screen_comparison_baf3_Imat_high_rep1vsrep2_duplex.csv",header = T,stringsAsFactors = F)[-1]
data$seqtech="duplex"
data_duplex=data


data=read.csv("Figures/Figure_1/outputs/baf3_Imat_high_rep1vsrep2_ngs/screen_comparison_baf3_Imat_high_rep1vsrep2_ngs.csv",header = T,stringsAsFactors = F)[-1]
data$seqtech="ngs"
data_ngs=data

data=read.csv("Figures/Figure_1/outputs/baf3_Imat_high_rep1vsrep2_sscs/screen_comparison_baf3_Imat_high_rep1vsrep2_sscs.csv",header = T,stringsAsFactors = F)[-1]
data$seqtech="sscs"
data_sscs=data

data=read.csv("Figures/Figure_1/outputs/baf3_Imat_high_rep1vsrep2_tileseq/screen_comparison_baf3_Imat_high_rep1vsrep2_tileseq.csv",header = T,stringsAsFactors = F)[-1]
data$seqtech="tileseq"
data_tileseq=data

data=rbind(data_duplex,data_ngs,data_sscs,data_tileseq)

data_sum=data%>%
  filter(protein_start%in%c(242:322),is_intended%in%1)%>%
  group_by(seqtech)%>%
  summarize(mean_depth=mean(depth_screen1_before),
            num=n())

data=data%>%rowwise%>%mutate(n_nuc=nchar(alt))

```

### Signal to noise ratios
#### Plotting unintended vs intended mutant distributions across sequencing technologies. 
```{r}
ggplot(data%>%filter(ct_screen1_before>=5,seqtech%in%"ngs",n_nuc%in%1),aes(x=ct_screen1_before,fill=factor(is_intended,levels=c(1,0))))+geom_density(alpha=0.7)+scale_x_continuous("Number of times a variant was seen",trans="log10")+cleanup+scale_fill_manual(values = c("blue","red"))+scale_y_continuous("Density")+theme(legend.position = "none")
ggsave("Figures/Figure_1/fig1A_abl_ngs_untreated.pdf",width=3.5,height=2.5,units="in",useDingbats=F)

ggplot(data%>%filter(ct_screen1_before>=5,seqtech%in%"tileseq",n_nuc%in%1),aes(x=ct_screen1_before,fill=factor(is_intended,levels=c(1,0))))+geom_density(alpha=0.7)+scale_x_continuous("Number of times a variant was seen",trans="log10")+cleanup+scale_fill_manual(values = c("blue","red"))+scale_y_continuous("Density")+theme(legend.position = "none")
ggsave("Figures/Figure_1/fig1E_abl_tileseq_untreated.pdf",width=3.5,height=2.5,units="in",useDingbats=F)

ggplot(data%>%filter(seqtech%in%"duplex",n_nuc%in%1),aes(x=ct_screen1_before,fill=factor(is_intended,levels=c(1,0))))+geom_density(alpha=0.7)+scale_x_continuous("Number of times a variant was seen",trans="log10")+cleanup+scale_fill_manual(values = c("blue","red"))+scale_y_continuous("Density")+theme(legend.position = "none")
ggsave("Figures/Figure_1/fig1F_abl_duplex_untreated.pdf",width=3.5,height=2.5,units="in",useDingbats=F)



ggplot(data %>%filter(ct_screen1_before>=5,seqtech%in%"ngs",n_nuc%in%1),
       aes(x = ct_screen1_before, fill = factor(is_intended, levels = c(1,0)))) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 50) +
  scale_x_continuous("Number of times a variant was seen", trans = "log10") +
  cleanup +
  scale_fill_manual(values = c("blue", "red")) +
  ylab("Count") +
  theme(legend.position = "none")

ggplot(data %>%filter(ct_screen1_before>=5,seqtech%in%"tileseq",n_nuc%in%1),
       aes(x = ct_screen1_before, fill = factor(is_intended, levels = c(1,0)))) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 50) +
  scale_x_continuous("Number of times a variant was seen", trans = "log10") +
  cleanup +
  scale_fill_manual(values = c("blue", "red")) +
  ylab("Count") +
  theme(legend.position = "none")

ggplot(data %>%filter(seqtech%in%"duplex",n_nuc%in%1),
       aes(x = ct_screen1_before, fill = factor(is_intended, levels = c(1,0)))) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 50) +
  # scale_x_continuous("Number of times a variant was seen", trans = "log10") +
  cleanup +
  scale_fill_manual(values = c("blue", "red")) +
  ylab("Count") +
  theme(legend.position = "none")
```



### Unintended mutants coverage
```{r}
library(tidyr)
data_snr=data%>%filter(protein_start%in%c(242:322))%>%group_by(seqtech,is_intended,protein_start,n_nuc_min)%>%summarize(nmuts=n(),coverage_mean=mean(ct_screen1_before),coverage_sum=sum(ct_screen1_before),depth_mean=mean(depth_screen1_before))

data_snr_cast <- data_snr %>%
  pivot_wider(
    names_from = is_intended,
    values_from = c(nmuts,coverage_mean,coverage_sum,depth_mean),
    names_glue = "{.value}_{ifelse(is_intended == 1, 'intended', 'unintended')}"
  )
# data_snr_cast=data_snr_cast%>%mutate(snr=coverage_mean_intended/coverage_mean_unintended)
data_snr_cast=data_snr_cast%>%mutate(snr=coverage_sum_intended/coverage_sum_unintended)

# Demonstrating noise:
ggplot(data_snr_cast%>%filter(seqtech%in%c("duplex","tileseq")),aes(x=protein_start,y=coverage_sum_unintended))+
  geom_point(color="black",size=2,shape=21,aes(fill=seqtech))+
  # facet_wrap(~seqtech)+
  scale_fill_manual(values = c("#FB8072","#8ED3C7"))+
  scale_y_continuous("Coverage of unintended mutants",trans = "log10")+
  scale_x_continuous("Rediue on BCRABL")+
  cleanup+
  theme(legend.position = "none",
        legend.title = element_text(size=8),
        legend.text = element_text(size=8),
        axis.text.x=element_text(color="black",size=8),
        # axis.title.x=element_blank(),
        # axis.title.y=element_blank(),
        axis.text.y=element_text(color="black",size=8),
        plot.margin=grid::unit(c(1,1,1,1), "mm"))
# ggsave("output/errorrates_figures/plot1.pdf",width=2.5,height=2.5,units = "in",useDingbats=F)

ggplot(data_snr_cast%>%filter(seqtech%in%c("duplex","tileseq")),aes(x=protein_start,y=nmuts_unintended,fill=seqtech))+
  geom_point(color="black",size=2,shape=21,aes(fill=seqtech))+
  scale_fill_manual(values = c("#FB8072","#8ED3C7"))+
  geom_hline(yintercept = 40,linetype="dashed")+
  cleanup+
  theme(legend.position = "none",
        legend.title = element_text(size=8),
        legend.text = element_text(size=8),
        axis.text.x=element_text(color="black",size=8),
        # axis.title.x=element_blank(),
        # axis.title.y=element_blank(),
        axis.text.y=element_text(color="black",size=8),
        plot.margin=grid::unit(c(1,1,1,1), "mm"))
# ggsave("output/errorrates_figures/plot2.pdf",width=2.5,height=2.5,units = "in",useDingbats=F)
```
### Coverage of intended resistant mutants
```{r}
# Demonstrating coverage:
ggplot(data%>%filter(resmuts%in%T,is_intended%in%1,seqtech%in%c("duplex","tileseq")),aes(x=species,y=ct_screen1_before,fill=seqtech))+
  geom_point(color="black",shape=21)+
  geom_line()+
  scale_y_continuous("Day 0 Coverage of Resistant Mutant",trans="log10",limits=c(1,1.2e5))+
  cleanup+
  theme(axis.text.x=element_text(color="black",size=8,angle=90,hjust=.5,vjust=.5),
        legend.position = "none",
        legend.title = element_text(size=8),
        legend.text = element_text(size=8),
        # axis.title.x=element_blank(),
        # axis.title.y=element_blank(),
        axis.text.y=element_text(color="black",size=8),
        plot.margin=grid::unit(c(1,1,1,1), "mm"))
# ggsave("output/errorrates_figures/plot3.pdf",width=2.5,height=2.5,units = "in",useDingbats=F)

ggplot(data%>%filter(resmuts%in%T,is_intended%in%1,seqtech%in%c("duplex","tileseq")),aes(x=species,y=ct_screen2_after,fill=seqtech))+
  geom_point(color="black",shape=21)+
  geom_line()+
  scale_y_continuous("Day 6 Coverage of Resistant Mutant",trans="log10",limits=c(1,1.2e5))+
  cleanup+
  theme(axis.text.x=element_text(color="black",size=8,angle=90,hjust=.5,vjust=.5),
        legend.position = "none",
        legend.title = element_text(size=8),
        legend.text = element_text(size=8),
        # axis.title.x=element_blank(),
        # axis.title.y=element_blank(),
        axis.text.y=element_text(color="black",size=8),
        plot.margin=grid::unit(c(1,1,1,1), "mm"))
# ggsave("output/errorrates_figures/plot3b.pdf",width=2.5,height=2.5,units = "in",useDingbats=F)
```

### CDF across sequencing technologies
```{r}
ggplot(data%>%filter(is_intended%in%1,seqtech%in%c("duplex","tileseq")),aes(x=ct_screen1_before,color=seqtech))+
  stat_ecdf()+
  scale_x_continuous("Intended mutant coverage",trans="log10")+
  scale_y_continuous("CDF")+
  scale_color_manual(values=c("#FB8072","#8ED3C7"))+
  cleanup+
  theme(axis.text.x=element_text(color="black",size=8,angle=90,hjust=.5,vjust=.5),
        legend.position = "none",
        legend.title = element_text(size=8),
        legend.text = element_text(size=8),
        # axis.title.x=element_blank(),
        # axis.title.y=element_blank(),
        axis.text.y=element_text(color="black",size=8),
        plot.margin=grid::unit(c(1,1,1,1), "mm"))
# ggsave("output/errorrates_figures/plot4.pdf",width=2.5,height=2.5,units = "in",useDingbats=F)


```

### Unintended mutants coverage more details
#### The big caveat when comparing the depths of NGS vs Tileseq data is that the TileSeq data was given approximately 10 times more sequencing reads. Therefore, at face value, it looks like TileSeq detects as many errors as NGS, even though that is not the case.
```{r}
ggplot(data_snr_cast%>%filter(seqtech%in%c("duplex","tileseq","ngs")),aes(x=protein_start,y=coverage_sum_unintended))+
  geom_point(color="black",size=2,shape=21,aes(fill=seqtech))+
  # facet_wrap(~seqtech)+
  scale_fill_manual(values = c("#FB8072","#3287BD","#8ED3C7"))+
  scale_y_continuous("Coverage of unintended mutants",trans = "log10")+
  scale_x_continuous("Rediue on BCRABL")+
  cleanup+
  theme(legend.position = "none",
        legend.title = element_text(size=8),
        legend.text = element_text(size=8),
        axis.text.x=element_text(color="black",size=8),
        # axis.title.x=element_blank(),
        # axis.title.y=element_blank(),
        axis.text.y=element_text(color="black",size=8),
        plot.margin=grid::unit(c(1,1,1,1), "mm"))

```

