---
title: "Mutation Resistance Map by Highest Pharmacological Dose"
author: "Marta Tomaszkiewicz"
date: "7/7/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set root directory to the current Rmd location
if (requireNamespace("rstudioapi", quietly = TRUE)) {
  knitr::opts_knit$set(root.dir = dirname(rstudioapi::getSourceEditorContext()$path))
}

library(ggplot2)
library(dplyr)
```

```{r load data}
data <- read.csv("all_mutants_with_relative_viability_at_three_clinical_doses.csv", 
                 header = TRUE, stringsAsFactors = FALSE)
```

```{r classify resistance}
# Define threshold
threshold_444 <- 0.420968716
threshold_760 <- 0.420968716
threshold_916 <- 0.420968716

# Resistance classification at 400 mg QD, 400 mg BID, and 500 mg BID
data$Resistant_444nM <- data$rel.viab.444 > threshold_444
data$Resistant_760nM <- data$rel.viab.760 > threshold_760
data$Resistant_916nM <- data$rel.viab.916 > threshold_916

# Assign highest resistant dose
data <- data %>%
  mutate(
    Highest_Resistant_Dose = case_when(
      is.na(rel.viab.444) & is.na(rel.viab.760) & is.na(rel.viab.916) ~ NA_character_,
      Resistant_916nM ~ "916nM",
      Resistant_760nM ~ "760nM",
      Resistant_444nM ~ "444nM",
      TRUE ~ "Sensitive"
    )
  )

# Define dose colors
dose_colors <- c(
  "444nM" = "#1f78b4",
  "760nM" = "#f46d43",
  "916nM" = "#a50026",
  "Sensitive" = "#f7f7f7"
)
```

```{r add highlights}
# Annotate highlight groups
data$highlight <- ifelse(
  data$species %in% c("T315I", "G250E", "E255K", "M351T", "Y253H", "M244V", "F317L",
                      "F359V", "H396R", "E255V", "E355G", "Q252H", "E459K", "L248V",
                      "Y253F", "D276G", "F359C", "L387M", "F359I", "F486S", "V379I",
                      "L384M", "F311L", "E279K", "H396P", "E355A", "M388L", "F317V",
                      "L387F", "S348L", "D325N", "E255D", "E275K", "F317C", "L248R", "T315V"),
  "highlight",
  ifelse(
    data$species %in% c("F311I", "Y353H", "K247R", "L298V", "E459G", "S420F", "T277A", 
                        "E282Q", "E352G", "E453K", "L273M", "L364I", "R332W", "S417Y", 
                        "T315A", "A366V", "A397P", "A407V", "E409K", "E450G", "F382L", 
                        "G321E", "M318I", "P441L", "P480L", "R473Q", "T277I", "V289F",
                        "W430L", "Y257C", "Y326H"), 
    "Sanger_VUDR", "normal"
  )
)
```

```{r plot resistance map}
# Create background grid
background_grid <- expand.grid(
  protein_start = seq(241, 513, by = 1),
  alt_aa = c("P", "G", "Y", "W", "F", "V", "L", "I", "A", "T", 
             "S", "Q", "N", "M", "C", "E", "D", "R", "K", "H")
)
background_grid$fill_value <- NA

# Merge background with data
merged_data <- merge(background_grid, data, 
                     by = c("protein_start", "alt_aa"), 
                     all.x = TRUE)

merged_data$fill_value <- ifelse(
  is.na(merged_data$Highest_Resistant_Dose),
  NA,
  merged_data$Highest_Resistant_Dose
)

# Add dummy rows for legend
create_dummy_row <- function(df, fill_value, highlight_value) {
  dummy <- df[1, ]
  dummy[,] <- NA
  dummy$fill_value <- fill_value
  dummy$highlight <- highlight_value
  return(dummy)
}

dummy_highlight <- create_dummy_row(merged_data, "highlight", "highlight")
dummy_sanger_vudr <- create_dummy_row(merged_data, "Sanger_VUDR", "Sanger_VUDR")
legend_data <- rbind(merged_data, dummy_highlight, dummy_sanger_vudr)

# Save data for reference
write.csv(legend_data, "Highest_resistance_mutation_classification.csv", row.names = FALSE)

# Define highlight border colors
highlight_colors <- c(
  highlight = "#33a02c",
  Sanger_VUDR = "#c51b7d"
)

# Create plot
plot <- ggplot(legend_data, aes(x = protein_start, y = alt_aa, fill = fill_value)) +
  geom_tile(width = 1, height = 1, na.rm = TRUE) +
  geom_tile(data = subset(merged_data, highlight == "highlight"),
            color = "#33a02c", size = 0.3, na.rm = TRUE) +
  geom_tile(data = subset(merged_data, highlight == "Sanger_VUDR"),
            color = "#c51b7d", size = 0.3, na.rm = TRUE) +
  scale_x_continuous(limits = c(242, 512), expand = c(0, 0)) +
  scale_fill_manual(
    values = c(dose_colors, highlight_colors),
    breaks = c("444nM", "760nM", "916nM", "Sensitive", "highlight", "Sanger_VUDR", NA),
    labels = c("400 mg QD", "400 mg BID", "500 mg BID", "Sensitive", "Clinical", "VUDR", "NA"),
    na.value = "grey80", na.translate = TRUE
  ) +
  guides(
    fill = guide_legend(
      direction = "horizontal",
      nrow = 1,
      override.aes = list(
        fill = c(dose_colors, "white", "white", "grey80"),
        color = c(rep("grey50", 4), "#33a02c", "#c51b7d", "grey50"),
        size = rep(4, 7)
      )
    )
  ) +
  labs(
    title = "Mutation Resistance Map by Highest Pharmacological Dose",
    x = "Residue position on BCR-ABL kinase",
    y = "Mutant Amino Acid",
    fill = "Highest Dose"
  ) +
  coord_fixed(ratio = 4) +
  theme_bw() +
  scale_y_discrete(
    limits = c("P", "G", "Y", "W", "F", "V", "L", "I", "A", "T", 
               "S", "Q", "N", "M", "C", "E", "D", "R", "K", "H")
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(color = "grey80"),
    axis.text.x = element_text(color = "black", face = "bold", size = 9),
    axis.text.y = element_text(color = "black", face = "bold", size = 9),
    aspect.ratio = 1/4,
    strip.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.key.size = unit(0.8, "lines"),
    legend.text = element_text(size = 9)
  )

print(plot)

ggsave("Map_by_highest_dose.pdf", plot, dpi = 600, width = 11, height = 8)
