---
title: "VUDR classification by highest clinical dose"
author: "Marta Tomaszkiewicz"
date: "2025-07-07"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set root directory to the current Rmd location
if (requireNamespace("rstudioapi", quietly = TRUE)) {
  knitr::opts_knit$set(root.dir = dirname(rstudioapi::getSourceEditorContext()$path))
}

library(ggplot2)
library(dplyr)
```


```{r load data}
data <- read.csv("VUDR_data_classified.csv", header = TRUE, stringsAsFactors = FALSE)
```


```{r classify resistance}
# Define custom colors
dose_colors <- c(
  "444nM" = "#1f78b4",
  "760nM" = "#f46d43",
  "916nM" = "#a50026",
  "Sensitive" = "#f7f7f7", 
  "NA" = "lightgrey"
)

# Calculate the count of each dose type
dose_counts <- data %>%
  group_by(Highest_Resistant_Dose) %>%
  summarise(count = n())

# First, create a named vector of labels with counts
dose_counts$legend_label <- paste0(dose_counts$Highest_Resistant_Dose, " (", dose_counts$count, ")")
names(dose_counts$legend_label) <- dose_counts$Highest_Resistant_Dose
```

```{r plot VUDR classification by highest resistance dose}
plot <- ggplot(dose_counts, aes(x = "", y = count, fill = Highest_Resistant_Dose)) +
  geom_bar(stat = "identity", width = 1, color = "black", linewidth = 0.3) +
  coord_polar(theta = "y", start = 0) +
  scale_fill_manual(values = dose_colors, labels = dose_counts$legend_label) +  # Add legend labels
  theme_void() +
  theme(legend.title = element_blank(),
        legend.position = "right") +
  ggtitle("Classification of VUDR by Highest Resistant Dose") +
  theme(plot.title = element_text(hjust = 0.5))

print (plot)

ggsave("VUDR classification by highest clinical dose.pdf", plot, dpi = 600, width = 8, height = 6, units = "in", bg = "white")

