---
title: "gnomad_analysis"
author: "Haider Inam"
date: '2025-07-24'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
# rm(list=ls())

library(stringr)
library(dplyr)
library(ggplot2)
library(plotly)
library(tictoc)
library(doParallel)
library(foreach)
library(RColorBrewer)
library(reshape2)

#Cleanup code for plotting
cleanup=theme_bw() +
  theme(plot.title = element_text(hjust=.5),
        panel.grid.major = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.background = element_blank(),
        # axis.line = element_line(color = "black"),
        axis.text = element_text(face="bold",color="black",size="11"),
        text=element_text(size=11,face="bold"),
        axis.title=element_text(face="bold",size="11"))
```

# Functions
## Function 1
```{r}
# Now that I have the ALT Codon from each gnomad codon, I am going to figure out the subset of amino acids that are possible at an n_nuc of 1 from the alt codon and from the ref codon.
codon_table=read.csv("Figures/Figure_5/codon_table.csv",header = T,stringsAsFactors = F)
# Inputs: ref codon, alt_codon (which really is the new ref codon)
# Outputs: list of amino acids possible at an n_nuc of 1 with the ref codon but not with the alt codon
# refseq_codon="AAA"
# gnomad_codon="ATA"
# refseq_codon="GAA"
# gnomad_codon="GAG"
# rm(ref_codon)
gnomad_unique_AAs=function(refseq_codon,gnomad_codon){
  refseq_codons=codon_table%>%
    # filter(Letter%in%alt_aa)%>%
    rowwise()%>%
    mutate(distance=as.numeric(!substr(refseq_codon,1,1)%in%substr(Codon,1,1))+
             as.numeric(!substr(refseq_codon,2,2)%in%substr(Codon,2,2))+
             as.numeric(!substr(refseq_codon,3,3)%in%substr(Codon,3,3)),
           subs_1=case_when(!substr(refseq_codon,1,1)%in%substr(Codon,1,1)~paste(substr(refseq_codon,1,1),">",substr(Codon,1,1),sep = ""),
                            T~""),
           subs_2=case_when(!substr(refseq_codon,2,2)%in%substr(Codon,2,2)~paste(substr(refseq_codon,2,2),">",substr(Codon,2,2),sep = ""),
                            T~""),
           subs_3=case_when(!substr(refseq_codon,3,3)%in%substr(Codon,3,3)~paste(substr(refseq_codon,3,3),">",substr(Codon,3,3),sep = ""),
                            T~""))%>%
    ungroup()

  refseq_codons=refseq_codons%>%filter(distance%in%1)
  # sort(unique(refseq_codons$Letter))
  
  gnomad_codons=codon_table%>%
      # filter(Letter%in%alt_aa)%>%
      rowwise()%>%
      mutate(distance=as.numeric(!substr(gnomad_codon,1,1)%in%substr(Codon,1,1))+
               as.numeric(!substr(gnomad_codon,2,2)%in%substr(Codon,2,2))+
               as.numeric(!substr(gnomad_codon,3,3)%in%substr(Codon,3,3)),
             subs_1=case_when(!substr(gnomad_codon,1,1)%in%substr(Codon,1,1)~paste(substr(gnomad_codon,1,1),">",substr(Codon,1,1),sep = ""),
                              T~""),
             subs_2=case_when(!substr(gnomad_codon,2,2)%in%substr(Codon,2,2)~paste(substr(gnomad_codon,2,2),">",substr(Codon,2,2),sep = ""),
                              T~""),
             subs_3=case_when(!substr(gnomad_codon,3,3)%in%substr(Codon,3,3)~paste(substr(gnomad_codon,3,3),">",substr(Codon,3,3),sep = ""),
                              T~""))%>%
      ungroup()
  
  gnomad_codons=gnomad_codons%>%filter(distance%in%1)
  gnomad_unique=sort(unique(gnomad_codons$Letter))
  # sort(unique(gnomad_codons$Letter))
  # Figuring out which gnomad amino acids are unique (i.e. not possible by refseq codon)
  gnomad_unique_norefseq=gnomad_unique[!sort(unique(gnomad_codons$Letter))%in%sort(unique(refseq_codons$Letter))]
  return(gnomad_unique_norefseq)
  
}


```


## Function 2
####Reading and processing the reference txt and csv file
```{r}
# #############For ABL#############
enst="ENST00000318560" #Ensembl transcript ID for the canonical transcript
chr=9 #The chromosome number on hg38. This argument is passed into the ensembl variant effect predictor
ref_txt_name="Figures/Figure_5/Refs/ABL/NM_005157.6_CDS.txt"
ref_csv_name="Figures/Figure_5/Refs/ABL/abl1_NM_005157.6_full_coordinates.csv"
mode="local"
ref_offset=0


#Reading the full sequence.
reference_seq=read.table(ref_txt_name)
reference_seq=as.character(reference_seq)

# substr(reference_seq,1,2)
#Ref_search is a function that returns the sequence of the cDNA, given start and stop coordinates
ref_search=function(start,stop){
  seachresult=substr(reference_seq,start,stop-1)
  seachresult
}

ref_genomic_coordinates=read.csv(ref_csv_name,header = T,stringsAsFactors = F)
# ref_genomic_coordinates=ref_genomic_coordinates%>%dplyr::select(-X)

####Adding sequence to each codon###
ref_genomic_coordinates=ref_genomic_coordinates%>%mutate(start=pos,end=start+3)%>%dplyr::select(!pos)
ref_genomic_coordinates=ref_genomic_coordinates%>%filter(!resi%in%NA)
ref_genomic_coordinates=ref_genomic_coordinates%>%
  rowwise()%>%
  mutate(codon=ref_search(start-ref_offset,end-ref_offset))
ref_genomic_coordinates=ref_genomic_coordinates%>%filter(!codon%in%"")
```

# Main gnomad analysis
## Data parsing
```{r}

gnomad=read.csv("Figures/Figure_5/gnomAD_v4.1.0_ENSG00000097007_2025_07_27_16_14_35.csv")
# 1100 Gnomad SNPS total
gnomad=gnomad%>%filter(Position>=130835447,Position<=130884388)
# Half of the SNPS are within the ABL coding sequence
gnomad=gnomad%>%filter(!VEP.Annotation%in%c("intron_variant","frameshift_variant"))
# Half of the SNPS in the kinase are non-intron non-synonymous
gnomad=gnomad%>%rowwise()%>%mutate(residue=substr(gsub("p\\.","",Protein.Consequence),4,6),
                         residue=as.numeric(gsub("([0-9]+).*$", "\\1", residue)))

gnomad=gnomad%>%filter(!residue%in%NA,residue<=516)

# Therefore, there are 343 gnomad variants in the in the SH3 or SH2 or Kinase domain of ABL
# 154 out of the 343 gnomad variants are in the kinase domain of ABL
# Within the gnomad variants in the kinase, only 10 are allele frequency > 1 in 100,000
# ggplot(gnomad,aes(x=Allele.Count))+stat_ecdf()+scale_x_continuous(trans="log10")

# There are 7 non-synonymous clinvar variants in the kinase domain in which there is a clinical annotation. Two of these are high frequency: R473Q/G (not provided clinvar significance), and K247R (benign). Both of these are neutral in our screen
# data=read.csv("output/ABLEnrichmentScreens/IL3_Enrichment_bgmerged_2.20.23.csv",header = T,stringsAsFactors = F)
# c=data%>%filter(protein_start%in%c(247,473))
#Of the clinvar variants 

# ABL kinase Spans from 130835447 to 130884388

gnomad=gnomad%>%
  rowwise()%>%
  mutate(residue=substr(gsub("p\\.","",Protein.Consequence),4,6),
         residue=as.numeric(gsub("([0-9]+).*$", "\\1", residue)),
         position=as.numeric(gsub("([0-9]+).*$", "\\1",strsplit(Transcript.Consequence,"\\.")[[1]][2])))

gnomad=gnomad%>%
  # rowwise()%>%
  mutate(ref=ref_search(position-ref_offset,position+1-ref_offset))

```
## Finding gnomad-enabled mutants
```{r}
#Aim is to create a list of gnomad enabled mutants, i.e. mutants that would have required multi-nucleotide variants before, but are now a single nucleotide away
# First: Make a function that calculates all possible amino acid substitutions for a given ref codon at a hamming distance of 1.
#A key pessimistic question to ask is: for any given residue, how many amino acid substitution types are only possible with two nucleotide substitutions? If basically every amino acid susbtitution is possible with a single nucleotide change, then it doesn't really matter if you make more possible with MNVs.
gnomad_simple=gnomad%>%dplyr::select(Position,Reference,Alternate,Protein.Consequence,residue,Transcript.Consequence,VEP.Annotation,Allele.Count,residue,position,ref)
# gnomad_simple=gnomad_simple%>%
#   rowwise()%>%
#   mutate(residue=substr(gsub("p\\.","",Protein.Consequence),4,6),
#          residue=as.numeric(gsub("([0-9]+).*$", "\\1", residue)),
#          position=as.numeric(gsub("([0-9]+).*$", "\\1",strsplit(Transcript.Consequence,"\\.")[[1]][2])))
# 
# gnomad_simple=gnomad_simple%>%
#   # rowwise()%>%
#   mutate(ref=ref_search(position-ref_offset,position+1-ref_offset))

#The following set of functions takes a position, ref, and and alt nucleotide, figures out which codon was mutated for each mutant, which index within the codon, and makes the alt_codon based on the new index 
# Alt offset: If alt_offset is 1, the mutation happens at position 1 in the codon, and so on
# Please note that a lot of this code relies on there being a single nucleotide substitution, i.e. it ignores mnvs and frameshifting indels. If frameshifts are included, then you have to think about variants that have long alt_offset lengths (not just 1 or 2 or 3).
# If its position 1: paste ALT with Second two Nucleotides
# If its position 2: paste first nucleotide, ALT nucleotide, and third nucleotide
# If it's position 3: paste second two nucleotides with the ALT nucleotide

gnomad=gnomad%>%
  rowwise()%>%
  mutate(ref_codon=(ref_genomic_coordinates[(position-ref_genomic_coordinates$start)%in%c(0,1,2),"codon"])[[1]],
         position_codon_end=(ref_genomic_coordinates[(position-ref_genomic_coordinates$start)%in%c(0,1,2),"end"])[[1]],
         alt_offset=position_codon_end-position,
         alt_codon=case_when(alt_offset%in%3~paste(Alternate,substr(ref_codon,2,3),sep=""),
                             alt_offset%in%1~paste(substr(ref_codon,1,2),Alternate,sep = ""),
                             alt_offset%in%2~paste(substr(ref_codon,1,1),Alternate,substr(ref_codon,3,3),sep = "")))
# ref_genomic_coordinates(130835462)
# ref_genomic_coordinates$start

gnomad=gnomad%>%dplyr::rename("refseq_codon"="ref_codon",
                                            "gnomad_codon"="alt_codon")

gnomad=gnomad%>%filter(nchar(gnomad_codon)<=3)

gnomad=gnomad%>%mutate(ref_aa=codon_table[refseq_codon==codon_table$Codon,"Letter"],
                                     gnomad_aa=codon_table[gnomad_codon==codon_table$Codon,"Letter"],
                                     species_enabler=paste(ref_aa,residue,gnomad_aa,sep = ""))

gnomad_simple=gnomad
########## Next figure out all unique gnomad codons by ABL residue.
# If a given residue has no gnomad SNPs, then there are no unique gnomad codons of course
#   What if a given position has two gnomad codons: figure out all possible gnomad codons and which ones are unique
# i=1
# gnomad_simple$gnomad_enabled_AAs=NA
for(i in seq(1:nrow(gnomad_simple))){
  gnomad_byiter=gnomad_simple[i,]
  enabler_i=gnomad_byiter$species_enabler
  resi_i=gnomad_byiter$residue
  enabled_i=gnomad_unique_AAs(gnomad_byiter$refseq_codon,gnomad_byiter$gnomad_codon)
  if(length(enabled_i)==0){
    # i.e. if there are no unique gnomad AAs, then skip to the next iteration in the loop
    next
  }
  gnomad_enabled_i=data.frame(enabler_i,enabled_i,resi_i)
  if(i==1){
    gnomad_enabled=gnomad_enabled_i
  }
  else if(!i==1){
    gnomad_enabled=rbind(gnomad_enabled,gnomad_enabled_i)
  }
}

gnomad_enabled=gnomad_enabled%>%distinct(enabler_i,enabled_i,resi_i)
gnomad_enabled=gnomad_enabled%>%dplyr::select(species_enabler=enabler_i,protein_start=resi_i,alt_aa=enabled_i)
gnomad_enabled$gnomad_enabled=T
# There are a total of 186 gnomad enabled ABL mutants. We see 72 of these mutants in our imatinib screens
# I have the gnomad enabled alternate amino acid by residue, but it might also be nice ot get a residue specific gnomad probability.
gnomad_pos_probabilities=gnomad%>%
  group_by(residue)%>%
  summarize(Allele.Count.Total=sum(Allele.Count),
                                         Allele.Number.Total=sum(Allele.Number))%>%
  mutate(Allele.Frequency.Gnomad=Allele.Count.Total/Allele.Number.Total)%>%
  # dplyr::select(-c("Allele.Count.Total","Allele.Number.Total"))%>%
  rename(protein_start=residue)

gnomad_enabled=merge(gnomad_enabled,gnomad_pos_probabilities,by=c("protein_start"),all.x = T)

# Addding ethnicity that each species was most overrepresented in
gnomad_enabled=merge(gnomad_enabled,gnomad%>%dplyr::select(species_enabler,GroupMax.FAF.group),by="species_enabler")

```
## Merging gnomad-enabled mutants with imatinib ABL DMS data
### This will enable analysis of the imatinib resistance status of these mutants

```{r}
# New Tileseq data as of 2.17.25
imatdata=read.csv("Figures/Figure_5/TileSeq_full_kinase_alldoses_with_lfc_corrected_netgrowths.csv",header = T,stringsAsFactors = F)
imatdata=imatdata%>%filter(ct_screen1_before>=5)

imatdata=imatdata%>%filter(is_intended%in%1,!(species=="R332M"& alt =="ATG"),!(species=="R460N"& alt =="AAC"))

#######Adding clinical prevalence predictions###############

resistance_predictions=read.csv("Figures/Figure_5/Tileseq_clinical_predictions_10.28.25.csv",header = T,stringsAsFactors = F)
resistance_predictions=resistance_predictions%>%dplyr::select(species,ref_aa,protein_start,alt_aa,fill_value,alpha.444,alpha.760,alpha.916)
resistance_predictions=resistance_predictions%>%filter(!ref_aa%in%NA)
imatdata=merge(imatdata,resistance_predictions,by=c("species","ref_aa","protein_start","alt_aa"))
###################################
imatdata=merge(imatdata,gnomad_enabled,by=c("protein_start","alt_aa"),all.x=T)
imatdata[imatdata$gnomad_enabled%in%NA,"gnomad_enabled"]=F

imatdata_gnomad=imatdata%>%filter(gnomad_enabled%in%T)%>%dplyr::select(dose,species,ref_aa,protein_start,alt_aa,netgr_obs_mean,netgr_obs_mean_corr,species_enabler,gnomad_enabled,GroupMax.FAF.group,Allele.Frequency.Gnomad,fill_value,alpha.444,alpha.760,alpha.916)


imatdata_gnomad=imatdata_gnomad%>%group_by(dose)%>%mutate(rank=rank(-netgr_obs_mean_corr))

library(forcats)
imatdata_gnomad=imatdata_gnomad%>%filter(!dose%in%"il3")
imatdata_gnomad <- imatdata_gnomad %>%
  mutate(species = fct_reorder(species, -rank, .desc = TRUE))

netgr_average=imatdata_gnomad%>%dplyr::select(dose,species_enabler,species,netgr_obs_mean_corr,rank)
netgr_average=netgr_average%>%filter(dose%in%"1200nM")

imatdata_gnomad$fill_value=factor(imatdata_gnomad$fill_value,
                                     levels=c("Sensitive",
                                              "444nM",
                                              "760nM",
                                              "916nM"))

```

### Updating dms dataset so that there are not multiple rows for each mutant type. This happens when two gnomad SNPs enable the same MNV. i.e. we don't want to double count enabled mutants.
```{r}

imatdata_gnomad <- imatdata_gnomad %>%
  group_by(species,dose) %>%
  summarise(
    # 1. First, create your custom aggregated column
    species_enabler = paste(unique(species_enabler), collapse = ", "),
    
    # 2. Now, apply 'first()' to all other columns
    # This line means: "For every column that is NOT 'species_enabler',
    # apply the 'first' function."
    # The 'species' column is the grouping variable, so it's not included here.
    across(!species_enabler, .fns = first)
  ) %>%
  ungroup() # Ungroup for good practice
```

## Generating supplemental table s13
```{r}
tables7=imatdata_gnomad%>%
  filter(dose%in%"1200nM")%>%
  dplyr::select(protein_start,
                alt_aa,
                species_enabler,
                GroupMax.FAF.group,
                species_enabler_allele.Frequency=Allele.Frequency.Gnomad,
                species_enabled=species,
                Highest_resistant_dose=fill_value)
# unique(tables7$GroupMax.FAF.group)
tables7=tables7%>%
  mutate(Overrepresented_grp=case_when(
    GroupMax.FAF.group%in%"nfe"~"European - non fin",
    GroupMax.FAF.group%in%"sas"~"South Asian",
    GroupMax.FAF.group%in%"amr"~"American",
    GroupMax.FAF.group%in%"mid"~"Middle Eastern",
    GroupMax.FAF.group%in%"afr"~"Afr/Afr American",
    GroupMax.FAF.group%in%"eas"~"East Asian",
    T~GroupMax.FAF.group
    ))

tables7=tables7%>%
  dplyr::select(protein_start,
                alt_aa,
                species_enabler,
                Overrepresented_grp,
                species_enabler_allele.Frequency,
                species_enabled,
                Highest_resistant_dose)
write.csv(tables7,"Figures/Figure_5/table_s13.csv")
```


## Data Plotting 

## Piecharts
Looking at which segments of the population are differentially affected by the gnomad mutants

```{r}
library(tidyverse)
gnomad <- gnomad %>%
  select(-starts_with("Homozygote"), -starts_with("Hemizygote"))

df_long <- gnomad %>%
  pivot_longer(
    cols = matches("Allele\\.(Count|Number)"),
    names_to = c(".value", "ancestry"),
    names_pattern = "Allele\\.(Count.|Number.)(.*)"
  )

df_long=df_long%>%filter(!ancestry%in%NA)%>%
  mutate(af=Count./Number.)
gnomad_enrichment_melt=df_long



library(tools)
gnomad_sum=gnomad_enrichment_melt%>%group_by(ancestry)%>%summarize(sum_freq=sum(af))
gnomad_sum=gnomad_sum%>%rowwise()%>%
  mutate(ancestry=gsub("\\."," ",ancestry),ancestry=toTitleCase(ancestry))
########################################################
# Piechart showing that the majority of individuals sequenced were of European Ancestry
# data=gnomad_enrichment_melt%>%
#   filter(residue%in%"315",rsIDs%in%"rs1338400730")
# 
# data <- data %>% 
#   arrange(desc(ancestry)) %>%
#   mutate(prop = Number. / sum(data$Number.) *100) %>%
#   mutate(ypos = cumsum(prop)- 0.5*prop )

#################################################################################



a=gnomad_enrichment_melt%>%filter(ancestry%in%"African.African.American")
nrow(a[!a$Count.==0,])
sum(a$Count.)

a=gnomad_enrichment_melt%>%filter(ancestry%in%"European..non.Finnish.")
nrow(a[!a$Count.==0,])
sum(a$Count.)
unique(gnomad_enrichment_melt$ancestry)
####################################################################################

gnomad_sum <- gnomad_sum %>% 
  arrange(desc(ancestry)) %>%
  mutate(prop = sum_freq / sum(gnomad_sum$sum_freq) *100) %>%
  mutate(ypos = cumsum(prop)- 0.5*prop )

data=gnomad_enrichment_melt%>%
  filter(!Count.%in%0)%>%
  group_by(ancestry)%>%
  summarize(nindividuals=sum(Count.),nmuts=n())

data <- data %>% 
  arrange(desc(nmuts)) %>%
  mutate(prop = nmuts / sum(data$nmuts) *100) %>%
  mutate(ypos = cumsum(prop)- 0.5*prop )




gnomad=gnomad%>%mutate(Allele.Count.Sum=case_when(Allele.Count%in%1~"1",))
nrow(gnomad[gnomad[,"Allele.Count"]==1,]) #Number of people with mutants only seen once
nrow(gnomad[gnomad[,"Allele.Count"]==2,]) #Number of people with mutants seen twice
sum(gnomad[gnomad[,"Allele.Count"]>=2,"Allele.Count"]) #Number of people with mutants seen more than twice
1614186-sum(gnomad[gnomad[,"Allele.Count"]>=2,"Allele.Count"]) #Number of people with mutants never seen in ABL
# There are 1599410 alleles in most ABL rows in gnomad data. This means that 1599410/2 = 799705 people were seen with mutations in ABL.
# Amongst these ~800k individuals, roughly SNPs were seen in roughly 147469, or 18.4% individuals


gnomad_piechart=data.frame(
  Allele.Count=c("0",">1"),
  Allele.Number=c(147469,652236)
)

ggplot(gnomad_piechart, aes(x="", y=Allele.Number,fill=Allele.Count)) +
  # geom_bar(stat="identity", width=1, color="white") +
    geom_col(color = "black") +
  coord_polar("y", start=2.3) +
  theme_void() + 
  theme(legend.position="none")+
  scale_fill_manual(values = c("#3953A4","gray70"))
# ggsave("Figures/Figure_5/gnomad_piechart_overall_percentage.pdf",width=1.5,height = 1.5, units="in",useDingbats=F)

ggplot(data, aes(x="", y=nindividuals, fill=ancestry))+
  geom_col()+
  theme_void() + 
  theme(legend.position="none") +
  geom_text(aes(label = ancestry),
            position = position_stack(vjust = 0.5),size=3.2) +
  # geom_text(aes(y = ypos, label = round(sum_freq,2)), color = "black", size=4) +
  scale_fill_brewer(palette="Set3")

# ggsave("output/gnomad_figures/gnomad_col_individuals.pdf",width=1,height = 2, units="in",useDingbats=F)

ggplot(data, aes(x="", y=nmuts, fill=ancestry))+
  geom_col()+
  theme_void() + 
  theme(legend.position="none") +
  geom_text(aes(label = ancestry),
            position = position_stack(vjust = 0.5),size=3.2) +
  # geom_text(aes(y = ypos, label = round(sum_freq,2)), color = "black", size=4) +
  scale_fill_brewer(palette="Set3")
# ggsave("output/gnomad_figures/gnomad_col_mutants.pdf",width=1,height = 2, units="in",useDingbats=F)


```

## Plotting the fitness distribution of gnomad-enabled mutants, colored by the ethnicities  

```{r}
imatdata_gnomad <- imatdata_gnomad %>%
  mutate(species = fct_reorder(species, -rank, .desc = TRUE))
# imatdata_gnomad=imatdata_gnomad%>%arrange(desc(rank))
library(tidyverse)
# quartz(type = 'pdf', file ='output/gnomad_figures/gnomad_enabled_mutants_penetrance.pdf',width=7,height = 4)

netgr_average=imatdata_gnomad%>%dplyr::select(dose,species_enabler,species,netgr_obs_mean_corr,rank,GroupMax.FAF.group)
netgr_average=netgr_average%>%filter(dose%in%"1200nM",!GroupMax.FAF.group%in%"",rank<400)
# a=imatdata_gnomad%>%filter(dose%in%"1200nM",!GroupMax.FAF.group%in%"",rank<400)
# quartz(type = 'pdf', file ='output/gnomad_figures/gnomad_enabled_mutants_penetrance.pdf',width=7,height = 4)
# a=imatdata_gnomad%>%filter(dose%in%"1200nM",!GroupMax.FAF.group%in%"")
ggplot(imatdata_gnomad%>%filter(dose%in%"1200nM",!GroupMax.FAF.group%in%"",rank<400)%>%arrange(desc(rank)),aes(x=reorder(species, -netgr_obs_mean_corr),y=netgr_obs_mean_corr,label=species,size=Allele.Frequency.Gnomad))+
  geom_point(aes(color=GroupMax.FAF.group))+
  geom_text(data = netgr_average, aes(x = species, y = netgr_obs_mean_corr, label = species),
            position = position_nudge(y = 0.032), # Adjust the position if needed
            size = 1.5, color = "red",
            angle=90)+
    geom_text(aes(label = "\u2191"), size = 4,position = position_nudge(y = 0.022)) +  # Add arrow shape
  geom_text(data = netgr_average, aes(x = species, y = netgr_obs_mean_corr, label = species_enabler),
            position = position_nudge(y = 0.012), # Adjust the position if needed
            size = 1.5, color = "blue",
            angle=90)+
  scale_x_discrete("SNP-enabled ABL mutant")+
  scale_y_continuous("Net growth rate (Hrs ^-1)",limits=c(-0.055,0.075))+
  scale_color_manual(values = c("#FFFFB3", #african,
                               "#8DD3C7", #American
                               "#FB8072", #East Asian
                               "#FCCDE5", #Middle Eastern
                               "#B3DE69", #NFE
                               "#BC80BD"))+#SAS
  cleanup+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0.5))

ggsave("Figures/Figure_5/gnomad_enabled_mutants_penetrance2_withlegend.pdf",width=7,height = 4, units="in",useDingbats=F)


netgr_average=imatdata_gnomad%>%dplyr::select(dose,species_enabler,species,netgr_obs_mean_corr,rank,GroupMax.FAF.group)
netgr_average=netgr_average%>%filter(dose%in%"1200nM")

ggplot(imatdata_gnomad%>%filter(dose%in%"1200nM")%>%arrange(desc(rank)),aes(x=reorder(species, -netgr_obs_mean_corr),y=netgr_obs_mean_corr,label=species,size=Allele.Frequency.Gnomad))+
  geom_point(aes(color=GroupMax.FAF.group))+
  # geom_text(data = netgr_average, aes(x = species, y = netgr_obs_mean_corr, label = species),
  #           position = position_nudge(y = 0.032), # Adjust the position if needed
  #           size = 1.5, color = "red",
  #           angle=90)+
  #   geom_text(aes(label = "\u2191"), size = 4,position = position_nudge(y = 0.022)) +  # Add arrow shape
  # geom_text(data = netgr_average, aes(x = species, y = netgr_obs_mean_corr, label = species_enabler),
  #           position = position_nudge(y = 0.012), # Adjust the position if needed
  #           size = 1.5, color = "blue",
  #           angle=90)+
  scale_x_discrete("SNP-enabled ABL mutant")+
  scale_y_continuous("Net growth rate (Hrs ^-1)",limits=c(-0.06,0.075))+
  scale_color_manual(values = c("#D9D9D9", #Other
                                "#FFFFB3", #african,
                               "#8DD3C7", #American
                               "#FB8072", #East Asian
                               "#FCCDE5", #Middle Eastern
                               # "#FDB462", #FE
                               "#B3DE69", #NFE
                               "#BC80BD"))+#SAS
  cleanup+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0.5,size=7),
        axis.title.x = element_text(size=9),
        axis.title.y = element_text(size=9))
ggsave("Figures/Figure_5/gnomad_enabled_mutants_penetrance.pdf",width=6,height = 3, units="in",useDingbats=F)

netgr_average=imatdata_gnomad%>%dplyr::select(dose,species_enabler,species,netgr_obs_mean_corr,rank,GroupMax.FAF.group)
netgr_average=netgr_average%>%filter(dose%in%"1200nM",rank<50)

# quartz(type = 'pdf', file ='output/gnomad_figures/gnomad_enabled_mutants_penetrance_zoomed.pdf',width=3.5,height = 3)
ggplot(imatdata_gnomad%>%filter(dose%in%"1200nM",rank<50)%>%arrange(desc(rank)),aes(x=reorder(species, -netgr_obs_mean_corr),y=netgr_obs_mean_corr,label=species,size=Allele.Frequency.Gnomad))+
  # geom_point(aes(color=fill_value))+
  geom_point(aes(color=GroupMax.FAF.group))+
  geom_text(data = netgr_average, aes(x = species, y = netgr_obs_mean_corr, label = species),
            position = position_nudge(y = -0.005), # Adjust the position if needed
            size = 2, color = "red",
            angle=90)+
    geom_text(aes(label = "\u2191"), size = 4,position = position_nudge(y = -0.01)) +  # Add arrow shape
  geom_text(data = netgr_average, aes(x = species, y = netgr_obs_mean_corr, label = species_enabler),
            position = position_nudge(y = -0.015), # Adjust the position if needed
            size = 2, color = "blue",
            angle=90)+
  scale_x_discrete("SNP-enabled ABL mutant")+
  scale_y_continuous("Net growth rate (Hrs ^-1)")+
  scale_color_manual(values = c("#D9D9D9", #Other
                                # "#FFFFB3", #african,
                                # "#8DD3C7", #American
                                "#FB8072", #East Asian
                                # "#FCCDE5", #Middle Eastern
                                # "#FDB462", #FE
                                "#B3DE69", #NFE
                                "#BC80BD"))+#SAS
  cleanup+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0.5,size=7),
        axis.title.x = element_text(size=9),
        axis.title.y = element_text(size=9))
ggsave("Figures/Figure_5/gnomad_enabled_mutants_penetrance_zoomed.pdf",width=3.5,height = 2.5, units="in",useDingbats=F)

```

## Plotting the predicted imatinib dose that will sensitize these gnomad-enabled mutants in the clinic  

```{r}
imatdata_gnomad=imatdata_gnomad%>%mutate(netgr.444=0.055-alpha.444,
                                         netgr.760=0.055-alpha.760,
                                         netgr.916=0.055-alpha.916)
imatdata_gnomad=imatdata_gnomad%>%group_by(dose)%>%mutate(rank_netgr760=rank(-netgr.760))
imatdata_gnomad=imatdata_gnomad%>%group_by(dose)%>%mutate(rank_alpha444=rank(alpha.444))

netgr_average=imatdata_gnomad%>%dplyr::select(dose,species_enabler,species,alpha.444,rank_alpha444,fill_value)
netgr_average=netgr_average%>%filter(dose%in%"1200nM",!fill_value%in%"Sensitive")


imatdata_gnomad <- imatdata_gnomad %>%
  mutate(species = fct_reorder(species, -rank_alpha444, .desc = TRUE))

# Filtering for which gnomad-enabled mutants were resistant in imatinib dataset (73 mutants)
a=imatdata_gnomad%>%filter(dose%in%"1200nM")
# For these 73 mutants, looking at the gnomad allele fractions
b=gnomad_enrichment_melt%>%filter(species_enabler%in%unique(a$species_enabler))
# For these 73 mutants, filtering for mutants that are overrepresented in a specific population
c=b%>%filter(!GroupMax.FAF.group%in%"")

a=df_summary <- imatdata_gnomad%>%
  filter(dose%in%"1200nM",!fill_value%in%"Sensitive")
source("code/res_residues_adder.R")
a=res_residues_adder(a)



df_summary <- imatdata_gnomad%>%
  filter(dose%in%"1200nM") %>%
  group_by(fill_value) %>%
  summarise(count = n()) %>%
  mutate(percent = count / sum(count) * 100,
         label = paste0(fill_value, " (", round(percent, 1), "%)"))


# Assign a grouping for large text labels
df_summary <- df_summary %>%
  mutate(group_label = ifelse(fill_value == "Sensitive", "Sensitive", "Resistant"))

# Pie chart
ggplot(df_summary, aes(x = "", y = percent, fill = fill_value)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  theme_void() +
  
  # Legend for slice colors
  scale_fill_manual(values = c(
    "Sensitive" = "#F6F7F7",  # leave big white wedge
    "444nM" = "#0073B2",  # blue
    "760nM" = "#F16635",  # orange
    "916nM" = "#A11B1F"  # red
  ))+
  theme(legend.title = element_blank())

ggsave("Figures/Figure_5/clin_var_piechart.pdf",width=3.5,height = 2.5, units="in",useDingbats=F)


```

## Plotting the Fitness distribution of known imatinibR mutants vs the rest of the library
```{r}
imatdata=read.csv("Figures/Figure_5/TileSeq_full_kinase_alldoses_with_lfc_corrected_netgrowths.csv",header = T,stringsAsFactors = F)
imatdata=imatdata%>%filter(ct_screen1_before>=5)

source("code/resmuts_adder.R")
imatdata=resmuts_adder(imatdata)
a=imatdata%>%filter(resmuts%in%T,dose%in%"1200nM")
ggplot(imatdata%>%filter(dose%in%"1200nM",!protein_start%in%"359",!species%in%"V299L",is_intended%in%1),aes(y=netgr_obs_mean_corr,fill=resmuts))+
  geom_density(alpha=.7)+
  cleanup+
  scale_y_continuous(limits=c(-0.06,0.075))+
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),legend.position="none",
          panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),plot.background=element_blank())+
  scale_fill_manual(values = c("gray90","red"))
ggsave("Figures/Figure_5/resistance_distributions.pdf",width=1,height = 3, units="in",useDingbats=F)




```



