---
title: "Imatinib concentration response map"
author: "Marta Tomaszkiewicz"
date: "2025-07-07"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set root directory to the current Rmd location
if (requireNamespace("rstudioapi", quietly = TRUE)) {
  knitr::opts_knit$set(root.dir = dirname(rstudioapi::getSourceEditorContext()$path))
}

library(ggplot2)
library(dplyr)
```

```{r load data}
data <- read.csv("all_mutants_with_IC50_estimates.csv", 
                 header = TRUE, stringsAsFactors = FALSE)
```

```{r prepare IC50 bins}
# Create a background grid
background_grid <- expand.grid(
  protein_start = seq(241, 513, by = 1),
  alt_aa = c("P", "G", "Y", "W", "F", "V", "L", "I", "A", "T", 
             "S", "Q", "N", "M", "C", "E", "D", "R", "K", "H")
)

# Add placeholder column
background_grid$IC50 <- NA

# Merge with dataset
merged_data_IC50 <- merge(
  background_grid, 
  data[, c("protein_start", "alt_aa", "IC50")], 
  by = c("protein_start", "alt_aa"), 
  all.x = TRUE
)

merged_data_IC50 <- merge(
  background_grid, 
  data[, c("protein_start", "alt_aa", "IC50")], 
  by = c("protein_start", "alt_aa"), 
  all.x = TRUE
) %>%
  rename(IC50 = IC50.y) %>%  
  select(protein_start, alt_aa, IC50)  

# Fix netgrowth values
merged_data_IC50 <- merged_data_IC50 %>%
  mutate(IC50 = case_when(
    !is.na(IC50) & IC50 >= 2454.9258 ~ 2454.9258,
    !is.na(IC50) & IC50 <= 0 ~ 0,
    TRUE ~ IC50
  ))

# Define IC50 bins and corresponding colors from "Paired"
ic50_bins <- c(0, 300, 600, 1200, 2500)
paired_colors <- c("#ffffbf", "#2c7bb6", "#fdae61", "#d7191c")


# Convert IC50 into factors based on bins
merged_data_IC50$IC50_bin <- cut(merged_data_IC50$IC50, breaks = ic50_bins, include.lowest = TRUE)
```

```{r plot IC50s}
#Legend is in one row
plot <- ggplot(merged_data_IC50, aes(x = protein_start, y = alt_aa, fill = IC50_bin)) +
  geom_tile(width = 1, height = 1) + 
  scale_x_continuous(name = "Residue position on BCR-ABL kinase", limits = c(241, 513), expand = c(0, 0)) +
  
  scale_fill_manual(values = paired_colors, name = "IC50 Bins", na.value = "grey80") +
  
  guides(
    fill = guide_legend(
      direction = "horizontal",
      nrow = 1
    )
  ) +
  
  labs(
    title = "Mutation IC50 Map",
    x = "Residue position on BCR-ABL kinase",
    y = "Mutant Amino Acid"
  ) +
  
  theme_bw() + 
  scale_y_discrete(
    limits = c("P", "G", "Y", "W", "F", "V", "L", "I", "A", "T", 
               "S", "Q", "N", "M", "C", "E", "D", "R", "K", "H")
  ) +
  
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(color = "grey80"),
    aspect.ratio = 1/4,
    axis.text.x = element_text(color = "black", face = "bold", size = 9),
    axis.text.y = element_text(color = "black", face = "bold", size = 9),
    strip.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 0.5), 
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.key.size = unit(0.8, "lines"),
    legend.text = element_text(size = 9),
    plot.margin = grid::unit(c(1,1,1,1), "mm")
  )

print(plot)

ggsave("Concentration response map.pdf", plot, width = 11, height = 8, dpi = 600)
