---
title: "abl_Sanger_enrichment_analysis"
author: "Haider Inam"
date: '2026-08-26'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This analysis addresses Reviewer #1's question of whether DMS-predicted
resistance is supported by real-world clinical evidence. We restrict to the
variants reachable by a single nucleotide substitution (SNV-accessible),
cross-reference them against the Sanger/COSMIC CML mutation catalog, and test
whether variants predicted to be resistant are clinically observed more often
than variants predicted to be sensitive.

**Exclusion criterion.** All 36 previously annotated clinically resistant
variants are removed — both the 21 mutant standards used to fit the
DMS-to-single-mutant regression correction and to set the Youden-index
resistance threshold, and the 15 literature-curated resistant variants. The
analysis therefore contains no variant whose resistance was known before this
study. Of the 36, 32 are present in the SNV-accessible pool with a valid dose
call, leaving n = 1,504 variants.

Supplementary Figure 10 is a single panel: the percentage of variants in each
predicted dose class that have been observed in the clinical catalog.

```{r load-and-prep}
library(dplyr)
library(ggplot2)
library(stringr)

# -----------------------------------------------------------------------------
# Load data
# -----------------------------------------------------------------------------
dms_data     <- read.csv("Tileseq_clinical_predictions_10.28.25.csv")
sanger_data  <- read.csv("ABL_Sanger_Gene_mutations_2026.csv")
hamming_data <- read.csv("hammingdistance_ablmutants.csv")
known_res    <- read.csv("known_clinicallyresistant_mutations_betterannotations.csv")

# All previously annotated clinically resistant variants are excluded:
# the mutant standards (used to build the classifier) and the
# literature-curated resistant variants (used to test it).
excluded_muts <- unique(known_res$Collapsed)

cat("n annotated resistant variants excluded:", length(excluded_muts),
    "(", sum(grepl("standard", known_res$Mutation_type, ignore.case = TRUE)),
    "mutant standards +",
    sum(!grepl("standard", known_res$Mutation_type, ignore.case = TRUE)),
    "literature-curated )\n")

dose_levels <- c("Sensitive", "444nM", "760nM", "916nM")

# -----------------------------------------------------------------------------
# Variants reachable by a single nucleotide substitution (SNV-accessible)
# -----------------------------------------------------------------------------
valid_snvs <- hamming_data %>%
  group_by(species) %>%
  summarise(min_nuc_dist = min(n_nuc_min, na.rm = TRUE), .groups = "drop") %>%
  filter(min_nuc_dist <= 1) %>%
  pull(species)

# -----------------------------------------------------------------------------
# Clinical (Sanger) mutation counts; require >=2 patients to exclude likely
# sequencing artifacts
# -----------------------------------------------------------------------------
sanger_grouped <- sanger_data %>%
  mutate(Mutation = str_replace(AA.Mutation, "^p\\.", "")) %>%
  group_by(Mutation) %>%
  summarise(Total_Patients = sum(Count, na.rm = TRUE), .groups = "drop")

valid_sanger_muts <- sanger_grouped %>%
  filter(Total_Patients >= 2) %>%
  pull(Mutation)

# -----------------------------------------------------------------------------
# Analysis table: every SNV-accessible variant with a valid dose call that
# carries no previously established clinical resistance annotation
# -----------------------------------------------------------------------------
analysis_df <- dms_data %>%
  distinct(species, .keep_all = TRUE) %>%
  filter(species %in% valid_snvs,
         Highest_Resistant_Dose %in% dose_levels,
         !species %in% excluded_muts) %>%
  left_join(sanger_grouped, by = c("species" = "Mutation")) %>%
  mutate(
    Total_Patients     = ifelse(is.na(Total_Patients), 0, Total_Patients),
    Dose_Tier          = factor(Highest_Resistant_Dose, levels = dose_levels),
    Observed_in_Sanger = species %in% valid_sanger_muts,
    Prediction         = ifelse(Highest_Resistant_Dose == "Sensitive",
                                 "Predicted Sensitive", "Predicted Resistant")
  )

nrow(analysis_df)
table(analysis_df$Dose_Tier)
```

## Statistical tests

```{r stats}
# -----------------------------------------------------------------------------
# Chi-squared: are predicted-resistant variants over-represented among
# clinically observed mutations, relative to predicted-sensitive variants?
# -----------------------------------------------------------------------------
ct   <- table(analysis_df$Prediction, analysis_df$Observed_in_Sanger)
test <- chisq.test(ct, correct = FALSE)
print(ct)
print(test)

res_total  <- sum(analysis_df$Prediction == "Predicted Resistant")
res_obs    <- sum(analysis_df$Prediction == "Predicted Resistant" &
                    analysis_df$Observed_in_Sanger)
sens_total <- sum(analysis_df$Prediction == "Predicted Sensitive")
sens_obs   <- sum(analysis_df$Prediction == "Predicted Sensitive" &
                    analysis_df$Observed_in_Sanger)
fold       <- (res_obs / res_total) / (sens_obs / sens_total)

cat(sprintf("Predicted Resistant observed: %d / %d (%.1f%%)\n",
            res_obs, res_total, 100 * res_obs / res_total))
cat(sprintf("Predicted Sensitive observed: %d / %d (%.1f%%)\n",
            sens_obs, sens_total, 100 * sens_obs / sens_total))
cat(sprintf("Fold enrichment: %.2fx\n", fold))

# -----------------------------------------------------------------------------
# Cochran-Armitage trend across the four ordered dose classes: does the
# probability of clinical observation increase monotonically with predicted
# dose class? (This is the claim the figure makes.)
# -----------------------------------------------------------------------------
x    <- as.integer(analysis_df$Dose_Tier)
y    <- as.integer(analysis_df$Observed_in_Sanger)
pbar <- mean(y)
trend_z <- sum((x - mean(x)) * (y - pbar)) /
           sqrt(pbar * (1 - pbar) * sum((x - mean(x))^2))
trend_p <- 2 * pnorm(-abs(trend_z))
cat(sprintf("Cochran-Armitage trend: z = %.2f, p = %.3g\n", trend_z, trend_p))

# -----------------------------------------------------------------------------
# Per-class summary (the figure's underlying table)
# -----------------------------------------------------------------------------
tier_tab <- analysis_df %>%
  group_by(Dose_Tier) %>%
  summarise(Total_N = n(), Observed_N = sum(Observed_in_Sanger), .groups = "drop") %>%
  mutate(Percentage = 100 * Observed_N / Total_N)

stats_tab <- data.frame(
  N        = nrow(analysis_df),
  Res_Obs  = res_obs,  Res_N  = res_total,
  Sens_Obs = sens_obs, Sens_N = sens_total,
  Fold     = fold,
  ChiSq    = unname(test$statistic),
  ChiSq_p  = test$p.value,
  Trend_z  = trend_z,
  Trend_p  = trend_p
)

print(stats_tab, digits = 4)
print(as.data.frame(tier_tab), digits = 3)

write.csv(stats_tab, "sanger_enrichment_stats.csv", row.names = FALSE)
write.csv(tier_tab,  "sanger_enrichment_tiers.csv", row.names = FALSE)
```

## Supplementary Figure 10 (single panel)

```{r cosmic-figure, fig.width=5.5, fig.height=4.5}
# Manuscript-wide color scheme for dose classes
dose_cols <- c("Sensitive" = "#7fb0d3",
               "444nM"     = "#2166ac",
               "760nM"     = "#f4a582",
               "916nM"     = "#b2182b")

figure_cosmic <- ggplot(tier_tab, aes(Dose_Tier, Percentage, fill = Dose_Tier)) +
  geom_col(color = "black", width = 0.7) +
  geom_text(aes(label = paste0(sprintf("%.1f%%", Percentage),
                                "\n(", Observed_N, "/", Total_N, ")")),
            vjust = -0.3, size = 3.5, fontface = "bold") +
  annotate("text", x = 2.5, y = Inf,
           label = sprintf("%.1f-fold enrichment\nchi2 = %.2f, p = %s",
                           fold, test$statistic,
                           format(test$p.value, digits = 3, scientific = TRUE)),
           vjust = 1.3, size = 3.2, colour = "grey25") +
  scale_fill_manual(values = dose_cols, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.35))) +
  labs(x = "Predicted highest resistant dose",
       y = "% observed in Sanger\n(>= 2 patients)") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(face = "bold"),
        panel.grid.minor = element_blank())

ggsave("figure_cosmic_enrichment.pdf", figure_cosmic, width = 5.5, height = 4.5)
ggsave("figure_cosmic_enrichment.png", figure_cosmic, width = 5.5, height = 4.5, dpi = 300)

print(figure_cosmic)
```
