---
title: "Minimal number of concentrations required for VUDR classification"
author: "Marta Tomaszkiewicz"
date: "2025-07-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set root directory to the current Rmd location
if (requireNamespace("rstudioapi", quietly = TRUE)) {
  knitr::opts_knit$set(root.dir = dirname(rstudioapi::getSourceEditorContext()$path))
}

library(ggplot2)
library(scales)
library(dplyr)
```


```{r load data}
data <- read.csv("IC50_3_versus_10_concentrations.csv", 
                 header = TRUE, stringsAsFactors = FALSE)
```


```{r IC50 grouping}

data$group <- ifelse(
  data$IC50.3conc > 1.2,
  "high", "low"
)

# Subset for each group
high_group <- data %>%
  filter(group == "high")

low_group <- data %>%
  filter(group == "low")
```

```{r Pearson correlation}
# Compute Pearson correlations
cor_high <- cor.test(high_group$IC50.12conc, high_group$IC50.3conc, method = "pearson")
cor_low <- cor.test(low_group$IC50.12conc, low_group$IC50.3conc, method = "pearson")
```

```{r plot}

plot <- ggplot(data, aes(
  x = IC50.12conc,
  y = IC50.3conc
)) +
  geom_point(aes(color = group)) +
  scale_color_manual(values = c("high" = "#d7191c", "low" = "black")) +
  geom_hline(yintercept = 1.2, linetype = "dashed", color = "#d7191c") +
  geom_abline(linetype = "dashed") +
  scale_x_continuous(
    "IC50 (µM) from 10 data points",
    trans = pseudo_log_trans(sigma = 1),
    breaks = c(0.3, 0.6, 0.9, 1.2, 3, 5, 10),
    limits = c(0.1, 10)
  ) +
  scale_y_continuous(
    "IC50 (µM) from 3 data points",
    trans = pseudo_log_trans(sigma = 1),
    breaks = c(0.3, 0.6, 0.9, 1.2, 3, 5, 10),
    limits = c(0.1, 10)
  ) +
  # Add Pearson correlation annotations
  annotate("text", x = 0.15, y = 8.5, hjust = 0, color = "#d7191c",
           label = paste0("R (>1200 nM) = ", round(cor_high$estimate, 2), 
                          ", p = ", signif(cor_high$p.value, 2))) +
  annotate("text", x = 0.45, y = 0.2, hjust = 0, color = "black",
           label = paste0("R (<1200 nM) = ", round(cor_low$estimate, 2), 
                          ", p = ", signif(cor_low$p.value, 2))) +
  theme(legend.position = "none")

print(plot)

ggsave("IC50_correlation_3_versus_10_concentration.pdf", plot, width = 6, height = 6, units = "in")
