---
title: "GnomAD-Enabled Mutation Analysis"
author: "Haider Inam"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: cosmo
---

```{r setup, include=FALSE}
# --- Global Settings ---
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# --- Library Imports ---
library(here)         # For robust file paths
library(tidyverse)    # Includes dplyr, ggplot2, stringr
library(plotly)
library(tictoc)
library(doParallel)
library(RColorBrewer)

# --- Plotting Theme ---
# Consistent theme for all publication figures
theme_paper <- theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.background = element_blank(),
    axis.text = element_text(face = "bold", color = "black", size = 11),
    text = element_text(size = 11, face = "bold"),
    axis.title = element_text(face = "bold", size = 11)
  )

theme_set(theme_paper)

# --- Palette ---
ancestry_colors <- c(
  "nfe" = "#B3DE69",      # Non-Finnish European
  "afr" = "#FFFFB3",      # African/African American
  "amr" = "#8DD3C7",      # Latino/Admixed American
  "eas" = "#FB8072",      # East Asian
  "mid" = "#FCCDE5",      # Middle Eastern
  "sas" = "#BC80BD",      # South Asian
  "Other" = "#D9D9D9"
)
```


1. Functions and Helpers
1.1 Codon Accessibility Logic
Calculates amino acids accessible via a single nucleotide variant (SNV) from a reference codon vs. a gnomAD codon.
```{r}
# Load codon table globally for the function to access
# NOTE: Ensure this file exists at this path relative to your project root
codon_table <- read.csv(here("Figures/Figure_5/inputs/codon_table.csv"), header = TRUE, stringsAsFactors = FALSE)

get_unique_accessible_aas <- function(refseq_codon, gnomad_codon) {
  
  # Helper to find 1-step neighbors for a given codon
  get_neighbors <- function(target_codon) {
    codon_table %>%
      rowwise() %>%
      mutate(
        # Calculate Hamming distance (number of nucleotide differences)
        distance = sum(
          substr(target_codon, 1, 1) != substr(Codon, 1, 1),
          substr(target_codon, 2, 2) != substr(Codon, 2, 2),
          substr(target_codon, 3, 3) != substr(Codon, 3, 3)
        )
      ) %>%
      ungroup() %>%
      filter(distance == 1) %>%
      pull(Letter) %>%
      unique() %>%
      sort()
  }

  # Get set of AAs accessible from Reference
  ref_accessible <- get_neighbors(refseq_codon)
  
  # Get set of AAs accessible from Gnomad Variant
  gnomad_accessible <- get_neighbors(gnomad_codon)
  
  # Return AAs accessible ONLY via the gnomad variant (the "enabled" mutants)
  gnomad_unique <- setdiff(gnomad_accessible, ref_accessible)
  
  return(gnomad_unique)
}
```
1. Functions
1.1 Codon Logic Function
Logic to determine which amino acids are accessible from the GnomAD codon but not the Reference codon. (Logic preserved exactly from original gnomad_unique_AAs)
```{r}
# Load codon table
codon_table <- read.csv(here("Figures/Figure_5/inputs/codon_table.csv"), header = TRUE, stringsAsFactors = FALSE)

# Function: gnomad_unique_AAs
# Inputs: ref codon, gnomad codon (which acts as the new reference)
# Outputs: list of amino acids possible at an n_nuc of 1 with the gnomad codon but NOT the ref codon
get_unique_accessible_aas <- function(refseq_codon, gnomad_codon){
  
  # 1. Calculate neighbors for Reference Codon
  refseq_neighbors <- codon_table %>%
    rowwise() %>%
    mutate(
      distance = as.numeric(!substr(refseq_codon,1,1) %in% substr(Codon,1,1)) +
                 as.numeric(!substr(refseq_codon,2,2) %in% substr(Codon,2,2)) +
                 as.numeric(!substr(refseq_codon,3,3) %in% substr(Codon,3,3))
    ) %>%
    ungroup() %>%
    filter(distance == 1)
  
  # 2. Calculate neighbors for Gnomad Codon
  gnomad_neighbors <- codon_table %>%
    rowwise() %>%
    mutate(
      distance = as.numeric(!substr(gnomad_codon,1,1) %in% substr(Codon,1,1)) +
                 as.numeric(!substr(gnomad_codon,2,2) %in% substr(Codon,2,2)) +
                 as.numeric(!substr(gnomad_codon,3,3) %in% substr(Codon,3,3))
    ) %>%
    ungroup() %>%
    filter(distance == 1)
  
  # 3. Determine unique Gnomad Amino Acids
  gnomad_unique_letters <- sort(unique(gnomad_neighbors$Letter))
  ref_unique_letters    <- sort(unique(refseq_neighbors$Letter))
  
  # AAs present in Gnomad set but NOT in Reference set
  gnomad_unique_norefseq <- gnomad_unique_letters[!gnomad_unique_letters %in% ref_unique_letters]
  
  return(gnomad_unique_norefseq)
}
```



1.2 Reference Sequence Processing
Reading the reference files and mapping coordinates. (Logic preserved: Uses read.table and character subsetting)

```{r}
# --- Constants ---
ENST_ID <- "ENST00000318560" 
CHR <- 9 
REF_OFFSET <- 0

# --- Load Reference Sequence ---
# Note: Using read.table as per original logic to maintain strict consistency
ref_txt_path <- here("Figures/Figure_5/inputs/Refs/ABL/NM_005157.6_CDS.txt")
reference_seq_raw <- read.table(ref_txt_path)
reference_seq_string <- as.character(reference_seq_raw)

# Function: ref_search
# Returns cDNA sequence given start/stop coordinates
ref_search <- function(start, stop){
  # Original logic: substr uses 1-based indexing
  substr(reference_seq_string, start, stop - 1)
}

# --- Load Genomic Coordinates ---
ref_csv_path <- here("Figures/Figure_5/inputs/Refs/ABL/abl1_NM_005157.6_full_coordinates.csv")
ref_genomic_coordinates <- read.csv(ref_csv_path, header = TRUE, stringsAsFactors = FALSE)

# Add sequence to each codon mapping
ref_genomic_coordinates <- ref_genomic_coordinates %>%
  mutate(start = pos, end = start + 3) %>%
  select(-pos) %>%
  filter(!is.na(resi)) %>%
  rowwise() %>%
  mutate(codon = ref_search(start - REF_OFFSET, end - REF_OFFSET)) %>%
  ungroup() %>%
  filter(codon != "")
```

2. Data Parsing
Filtering GnomAD variants for ABL1 Kinase domain.

```{r}
gnomad_raw <- read.csv(here("Figures/Figure_5/inputs/gnomAD_v4.1.0_ENSG00000097007_2025_07_27_16_14_35.csv"))

# Filter for Kinase Domain and Coding Variants
gnomad_variants <- gnomad_raw %>%
  filter(Position >= 130835447, Position <= 130884388) %>%
  filter(!VEP.Annotation %in% c("intron_variant", "frameshift_variant")) %>%
  rowwise() %>%
  mutate(
    residue = substr(gsub("p\\.","", Protein.Consequence), 4, 6),
    residue = as.numeric(gsub("([0-9]+).*$", "\\1", residue))
  ) %>%
  ungroup() %>%
  filter(!is.na(residue), residue <= 516)

# Extract position information
gnomad_variants <- gnomad_variants %>%
  rowwise() %>%
  mutate(
    # Re-parsing residue to be safe (preserved from original)
    residue = substr(gsub("p\\.","", Protein.Consequence), 4, 6),
    residue = as.numeric(gsub("([0-9]+).*$", "\\1", residue)),
    
    # Parse Transcript Consequence for cDNA position
    position_cdna = as.numeric(gsub("([0-9]+).*$", "\\1", strsplit(Transcript.Consequence, "\\.")[[1]][2]))
  ) %>%
  ungroup() %>%
  mutate(
    ref_base = ref_search(position_cdna - REF_OFFSET, position_cdna + 1 - REF_OFFSET)
  )
```

3. Finding GnomAD-Enabled Mutants

```{r}
# --- Step 3.1: Construct Mutant Codons ---
gnomad_variants <- gnomad_variants %>%
  rowwise() %>%
  mutate(
    # Locate the reference codon frame
    ref_codon_seq = (ref_genomic_coordinates[(position_cdna - ref_genomic_coordinates$start) %in% c(0,1,2), "codon"])[[1]],
    codon_end_pos = (ref_genomic_coordinates[(position_cdna - ref_genomic_coordinates$start) %in% c(0,1,2), "end"])[[1]],
    
    # Determine offset within codon
    alt_offset = codon_end_pos - position_cdna,
    
    # Construct Alternate Codon
    alt_codon_seq = case_when(
      alt_offset == 3 ~ paste(Alternate, substr(ref_codon_seq, 2, 3), sep=""),
      alt_offset == 1 ~ paste(substr(ref_codon_seq, 1, 2), Alternate, sep=""),
      alt_offset == 2 ~ paste(substr(ref_codon_seq, 1, 1), Alternate, substr(ref_codon_seq, 3, 3), sep="")
    )
  ) %>%
  ungroup() %>%
  rename(refseq_codon = ref_codon_seq, gnomad_codon = alt_codon_seq) %>%
  filter(nchar(gnomad_codon) <= 3)

# Add Amino Acid Letters
gnomad_variants <- gnomad_variants %>%
  mutate(
    ref_aa = codon_table[match(refseq_codon, codon_table$Codon), "Letter"],
    gnomad_aa = codon_table[match(gnomad_codon, codon_table$Codon), "Letter"],
    species_enabler = paste(ref_aa, residue, gnomad_aa, sep = "")
  )

# --- Step 3.2: Iterate to find Enabled Mutants (The Loop) ---
# Preparing data for the loop
gnomad_simple <- gnomad_variants
gnomad_enabled_list <- list() # Store results in list for speed then bind

# Loop through every variant row
# NOTE: Preserving original loop structure
for(i in seq_len(nrow(gnomad_simple))){
  
  row_data <- gnomad_simple[i, ]
  enabler_id <- row_data$species_enabler
  residue_val <- row_data$residue
  
  # Call the helper function
  enabled_aas <- get_unique_accessible_aas(row_data$refseq_codon, row_data$gnomad_codon)
  
  # If new accessible AAs exist, store them
  if(length(enabled_aas) > 0){
    gnomad_enabled_list[[i]] <- data.frame(
      enabler_i = enabler_id, 
      enabled_i = enabled_aas, 
      resi_i = residue_val,
      stringsAsFactors = FALSE
    )
  }
}

# Combine results
gnomad_enabled <- do.call(rbind, gnomad_enabled_list) %>%
  distinct(enabler_i, enabled_i, resi_i) %>%
  select(species_enabler = enabler_i, protein_start = resi_i, alt_aa = enabled_i) %>%
  mutate(gnomad_enabled = TRUE)

# --- Step 3.3: Add Probabilities ---
# Calculate Allele Frequencies per residue
gnomad_pos_probabilities <- gnomad_variants %>%
  group_by(residue) %>%
  summarize(
    Allele.Count.Total = sum(Allele.Count),
    Allele.Number.Total = sum(Allele.Number)
  ) %>%
  mutate(Allele.Frequency.Gnomad = Allele.Count.Total / Allele.Number.Total) %>%
  rename(protein_start = residue)

# Merge back to enabled table
gnomad_enabled_final <- merge(gnomad_enabled, gnomad_pos_probabilities, by = "protein_start", all.x = TRUE)

# Add Ethnicity data
gnomad_enabled_final <- merge(
  gnomad_enabled_final, 
  gnomad_variants %>% select(species_enabler, GroupMax.FAF.group) %>% distinct(), 
  by = "species_enabler"
)
```

4. Merging with TileSeq Data
Integrating GnomAD findings with Imatinib resistance data.
```{r}
# Load Imatinib Data
imat_data <- read.csv(here("Figures/Figure_5/inputs/TileSeq_full_kinase_alldoses_with_lfc_corrected_netgrowths.csv"), stringsAsFactors = FALSE) %>%
  filter(ct_screen1_before >= 5) %>%
  filter(is_intended == 1, 
         !(species == "R332M" & alt == "ATG"), 
         !(species == "R460N" & alt == "AAC"))

# Load Clinical Predictions
res_predictions <- read.csv(here("Figures/Figure_5/inputs/Tileseq_clinical_predictions_10.28.25.csv"), stringsAsFactors = FALSE) %>%
  select(species, ref_aa, protein_start, alt_aa, fill_value, alpha.444, alpha.760, alpha.916) %>%
  filter(!is.na(ref_aa))

# Merge all datasets
imat_data <- merge(imat_data, res_predictions, by = c("species", "ref_aa", "protein_start", "alt_aa"))
imat_data <- merge(imat_data, gnomad_enabled_final, by = c("protein_start", "alt_aa"), all.x = TRUE)

# Mark non-enabled as False
imat_data$gnomad_enabled[is.na(imat_data$gnomad_enabled)] <- FALSE

# Create Final Subset for Plotting
imat_gnomad_subset <- imat_data %>%
  filter(gnomad_enabled == TRUE) %>%
  select(dose, species, ref_aa, protein_start, alt_aa, netgr_obs_mean, netgr_obs_mean_corr, 
         species_enabler, gnomad_enabled, GroupMax.FAF.group, Allele.Frequency.Gnomad, 
         fill_value, alpha.444, alpha.760, alpha.916)

# Ranking for Plot Order
imat_gnomad_subset <- imat_gnomad_subset %>%
  group_by(dose) %>%
  mutate(rank = rank(-netgr_obs_mean_corr)) %>%
  ungroup() %>%
  filter(dose != "il3") %>%
  mutate(species = fct_reorder(species, -rank, .desc = TRUE)) %>%
  mutate(fill_value = factor(fill_value, levels = c("Sensitive", "444nM", "760nM", "916nM")))

# --- Collapse Duplicates ---
# Merging rows where multiple SNPs enable the same mutant
imat_gnomad_subset <- imat_gnomad_subset %>%
  group_by(species, dose) %>%
  summarise(
    species_enabler = paste(unique(species_enabler), collapse = ", "),
    across(!species_enabler, first),
    .groups = "drop"
  )
```

5. Output Tables and Figures
5.1 Supplemental Table S13
```{r}
table_s13 <- imat_gnomad_subset %>%
  filter(dose == "1200nM") %>%
  mutate(Overrepresented_grp = case_when(
    GroupMax.FAF.group == "nfe" ~ "European - non fin",
    GroupMax.FAF.group == "sas" ~ "South Asian",
    GroupMax.FAF.group == "amr" ~ "American",
    GroupMax.FAF.group == "mid" ~ "Middle Eastern",
    GroupMax.FAF.group == "afr" ~ "Afr/Afr American",
    GroupMax.FAF.group == "eas" ~ "East Asian",
    TRUE ~ GroupMax.FAF.group
  )) %>%
  select(
    protein_start,
    alt_aa,
    species_enabler,
    Overrepresented_grp,
    species_enabler_allele.Frequency = Allele.Frequency.Gnomad,
    species_enabled = species,
    Highest_resistant_dose = fill_value
  )

write.csv(table_s13, here("Figures/Figure_5/table_s13.csv"), row.names = FALSE)
```


5.2 Ancestry Bar Charts
```{r}
# Prepare data (match original structure)
gnomad_ancestry_long <- gnomad_raw %>%
  select(-starts_with("Homozygote"), -starts_with("Hemizygote")) %>%
  pivot_longer(
    cols = matches("Allele\\.(Count|Number)"),
    names_to = c(".value", "ancestry"),
    names_pattern = "Allele\\.(Count.|Number.)(.*)"
  ) %>%
  filter(!is.na(ancestry)) %>%
  mutate(af = Count. / Number.)

# Summary for Plotting
ancestry_summary <- gnomad_ancestry_long %>%
  filter(Count. > 0) %>%
  group_by(ancestry) %>%
  summarize(
    nindividuals = sum(Count.), 
    nmuts = n()
  ) %>%
  mutate(
    ancestry_clean = tools::toTitleCase(gsub("\\.", " ", ancestry)),
    prop = nmuts / sum(nmuts) * 100
  ) %>%
  arrange(desc(nmuts))

# Plot: Mutants by Ancestry (Stacked Bar)
ggplot(ancestry_summary, aes(x = "", y = nmuts, fill = ancestry_clean)) +
  geom_col() +
  theme_void() + 
  theme(legend.position = "none") +
  geom_text(aes(label = ancestry_clean),
            position = position_stack(vjust = 0.5), size = 3.2) +
  scale_fill_brewer(palette = "Set3")

ggsave(here("Figures/Figure_5/gnomad_col_mutants.pdf"), width=1, height=2)

ggplot(ancestry_summary, aes(x = "", y = nindividuals, fill = ancestry_clean)) +
  geom_col() +
  theme_void() + 
  theme(legend.position = "none") +
  geom_text(aes(label = ancestry_clean),
            position = position_stack(vjust = 0.5), size = 3.2) +
  scale_fill_brewer(palette = "Set3")

ggsave(here("Figures/Figure_5/gnomad_col_individuals.pdf"), width=1, height=2)
```


## 5.3 Fitness Distribution (Dot Plot)
```{r fitness_plot}

# --- 1. Main Penetrance Plot (All Mutants) ---
plot_data <- imat_gnomad_subset %>%
  filter(dose == "1200nM") %>%
  arrange(desc(rank))

# Annotation subset for top ranked (same as original code)
labels_data <- plot_data %>% 
  filter(!GroupMax.FAF.group %in% "", rank < 400)

p_main <- ggplot(plot_data, 
       aes(x = reorder(species, -netgr_obs_mean_corr), 
           y = netgr_obs_mean_corr, 
           label = species, 
           size = Allele.Frequency.Gnomad)) +
  geom_point(aes(color = GroupMax.FAF.group)) +
  
  # Annotations for Main Plot
  # geom_text(data = labels_data, aes(label = species),
  #           nudge_y = 0.032, size = 1.5, color = "red", angle = 90) +
  # geom_text(data = labels_data, aes(label = "\u2191"), 
  #           nudge_y = 0.022, size = 3, color = "black") + 
  
  scale_x_discrete("SNP-enabled ABL mutant") +
  scale_y_continuous("Net growth rate (Hrs ^-1)", limits = c(-0.06, 0.075)) +
  scale_color_manual(values = ancestry_colors) +
  
  # Theme
  theme_paper +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none",
    axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 7),
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9)
  )

print(p_main)
ggsave(here("Figures/Figure_5/gnomad_enabled_mutants_penetrance.pdf"), p_main, width = 6, height = 3, useDingbats = FALSE)


# --- 2. Zoomed Penetrance Plot (Top 50 Ranks) ---
# Filter for top 50
zoomed_data <- imat_gnomad_subset %>%
  filter(dose == "1200nM", rank < 50) %>%
  arrange(desc(rank))

# Create annotation dataframe for zoomed plot
zoomed_annotations <- zoomed_data %>%
  select(dose, species_enabler, species, netgr_obs_mean_corr, rank, GroupMax.FAF.group)

zoomed_data=zoomed_data%>%
  mutate(GroupMax.FAF.group=case_when(GroupMax.FAF.group%in%""~"Other",
                                      T~GroupMax.FAF.group))

# quartz(type = 'pdf', file ='Figures/Figure_5/gnomad_enabled_mutants_penetrance_zoomed.pdf',width=3.5,height = 3)

ggplot(zoomed_data, 
       aes(x = reorder(species, -netgr_obs_mean_corr), 
           y = netgr_obs_mean_corr, 
           label = species, 
           size = Allele.Frequency.Gnomad)) +
  geom_point(aes(color = GroupMax.FAF.group)) +
  
  # Specific Annotations for Zoomed Plot (Nudged downwards)
  geom_text(data = zoomed_annotations, aes(label = species),
            position = position_nudge(y = -0.005), 
            size = 2, color = "red", angle = 90) +
  geom_text(aes(label = "\u2191"), size = 4, 
            position = position_nudge(y = -0.01)) +
  geom_text(data = zoomed_annotations, aes(label = species_enabler),
            position = position_nudge(y = -0.015), 
            size = 2, color = "blue", angle = 90) +
            
  scale_x_discrete("SNP-enabled ABL mutant") +
  scale_y_continuous("Net growth rate (Hrs ^-1)") +
  
  # Use consistent palette (handles the specific colors for East Asian/NFE/SAS/Other automatically)
  scale_color_manual(values = ancestry_colors) +
  
  theme_paper +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none",
    axis.text.y = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size = 7),
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9)
  )
# ggsave("Figures/Figure_5/gnomad_enabled_mutants_penetrance_zoomed.pdf",width=3.5,height = 2.5, units="in",useDingbats=F)
# print(p_zoomed)

```


5.4 Clinical Resistance (Pie Chart)
```{r}
res_summary_plot <- imat_gnomad_subset %>%
  filter(dose == "1200nM") %>%
  group_by(fill_value) %>%
  summarise(count = n()) %>%
  mutate(percent = count / sum(count) * 100)

ggplot(res_summary_plot, aes(x = "", y = percent, fill = fill_value)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  theme_void() +
  scale_fill_manual(values = c(
    "Sensitive" = "#F6F7F7",
    "444nM"     = "#0073B2",
    "760nM"     = "#F16635",
    "916nM"     = "#A11B1F"
  ))

ggsave(here("Figures/Figure_5/clin_var_piechart_cleaned.pdf"), width = 3.5, height = 2.5)
```


5.5 Fitness Distribution (Density)
Comparing resistant mutants vs library.
```{r}
# Source helper function if needed, or implement here
# source(here("code/resmuts_adder.R"))
# For now, assuming resmuts column exists or needs helper:
# Simple fallback if function missing:
imat_data$resmuts <- ifelse(imat_data$netgr_obs_mean_corr > 0.0, TRUE, FALSE) # Placeholder logic

ggplot(imat_data %>% filter(dose == "1200nM", is_intended == 1), 
       aes(y = netgr_obs_mean_corr, fill = resmuts)) +
  geom_density(alpha = 0.7) +
  scale_y_continuous(limits = c(-0.06, 0.075)) +
  scale_fill_manual(values = c("gray90", "red")) +
  theme_void() +
  theme(legend.position = "none")

ggsave(here("Figures/Figure_5/resistance_distributions_cleaned.pdf"), width = 1, height = 3)
```

6. Session Info
```{r}
sessionInfo()
```

