---
title: "dose_response_curve_fitting_with_errorbars"
author: "Haider Inam"
date: "6/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_knit$set(root.dir = normalizePath(".."))
```
Two ways of calculating errorbars for IC50 dose response data:  
Method 1. Directly calculate the mean, SD, and 95% CI of the 8 or so replicates for each mutant at each of the 10 doses for BCRABL Imatinib IC50s. One Caveat here: You can calculate the errorbars this way but the data still fits better if the confidence intervals are around a mean that is not the average of the replicates but the fit from the 4 parameter logistic. Sounds confusing, yes.
```{r}
library(knitr)
library(tictoc)
library(workflowr)
library(VennDiagram)
library(dplyr)
library(foreach)
library(doParallel)
library(ggplot2)
library(reshape2)
library(RColorBrewer)
library(devtools)
library(ggsignif)
library(plotly)
library(BiocManager)
library(drc)
#rm(list=ls())
net_gr_wodrug=0.055
conc_for_predictions=0.8

ic50data_all=read.csv("data/ic50data/IC50HeatMap.csv",header = T,stringsAsFactors = F)
# ic50data_all=read.csv("../data/IC50HeatMap.csv",header = T,stringsAsFactors = F)

ic50data_all=ic50data_all%>%filter(species%in%c("WT","V299L_H","E355A","D276G_maxi","H396R","F317L","F359I","E459K","G250E","F359C","F359V","M351T","L248V","E355G_maxi","Q252H_maxi","Y253F","F486S_maxi","H396P_maxi","E255K","Y253H","T315I","E255V"))


ic50data_all=ic50data_all%>%
  mutate(species=case_when(species=="F486S_maxi"~"F486S",
                          species=="H396P_maxi"~"H396P",
                          species=="Q252H_maxi"~"Q252H",
                          species=="E355G_maxi"~"E355G",
                          species=="D276G_maxi"~"D276G",
                          species=="V299L_H" ~ "V299L",
                          TRUE ~as.character(species)))

ic50data_long=read.csv("data/ic50data/ic50data_all_conc.csv",header = T,stringsAsFactors = F)
# ic50data_long=read.csv("../output/ic50data_all_conc.csv",header = T,stringsAsFactors = F)
ic50data_long$netgr_pred=net_gr_wodrug-ic50data_long$drug_effect


########################Getting Confidence Intervals########################
#will just start of with one dose at 1.25uM
ic50data_all2=data.frame(cbind(ic50data_all$species,ic50data_all$X1.25))
# ic50data_all2=data.frame(cbind(ic50data_all$species,ic50data_all$X0.625))
colnames(ic50data_all2)=c("species","dose_response")
ic50data_all2$dose_response=as.numeric(as.character(ic50data_all2$dose_response))


#Calculating confidence limit and standard deviations
ic50data_all_sum=ic50data_all2%>%group_by(species)%>%summarise(dr_mean=mean(dose_response),dr_ci_ul=dr_mean+1.96*sd(dose_response)*sqrt(n()),dr_ci_ll=dr_mean-1.96*sd(dose_response)*sqrt(n()),dr_sd_ul=dr_mean+sd(dose_response),dr_sd_ll=dr_mean-sd(dose_response))

#Since some mutants have 0% alive at the lower limit of the confidence interval, I'm converting those to 0
ic50data_all_sum=ic50data_all_sum%>%mutate(dr_ci_ll=case_when(dr_ci_ll<=0~0,
                                                              TRUE~dr_ci_ll),
                                           dr_sd_ll=case_when(dr_sd_ll<=0~0,
                                                              TRUE~dr_sd_ll))

ic50data_all_sum=merge(ic50data_all_sum,ic50data_long%>%filter(conc==conc_for_predictions)%>%dplyr::select(mutant,netgr_pred_model=netgr_pred),by.x="species",by.y="mutant")
#Dose response here is essentially y. aka %alive
#Converting y to drug effect on growth rate aka alpha value
ic50data_all_sum=ic50data_all_sum%>%
  mutate(netgr_pred_ci_ul=net_gr_wodrug-(-log(dr_ci_ul)/72),
         netgr_pred_mean=net_gr_wodrug-(-log(dr_mean)/72), #removing this because we want netgr_mean to be derived from the logistic fit
         netgr_pred_ci_ll=net_gr_wodrug-(-log(dr_ci_ll)/72),
         netgr_pred_sd_ul=net_gr_wodrug-(-log(dr_sd_ul)/72),
         netgr_pred_sd_ll=net_gr_wodrug-(-log(dr_sd_ll)/72),
         netgr_pred_model_sd_ul=netgr_pred_model+(netgr_pred_mean-netgr_pred_sd_ll),
         netgr_pred_model_sd_ll=netgr_pred_model-(netgr_pred_mean-netgr_pred_sd_ll))

####Since the more sensitive mutants were appearing to grow fast if I take the raw IC50 predicted growth rates, I am going to instead take the predicted growth rates from the IC50s that were fit on a 4-parameter logistic. To get standard deviations, I will just add/subtract the standard deviations from the regular plots.

# write.csv(ic50data_all_sum,"data/ic50data_all_confidence_intervals_raw_data.csv")

####Need to look into why method 1 gives growth rates with CIs that are evenly spread from the mean. I think this is because of this peculiar calculation netgr_pred_model_sd_ul=netgr_pred_model+(netgr_pred_mean-netgr_pred_sd_ll)
```




Method 2. Do a 4 parameter logistic fit on all the replicates individually. Then calculate the mean, standard deviations, and CIs off of the multiple logistic fits for each of the ~8ish replicates for each mutant.
```{r}
# rm(list=ls())
twinstrand_maf_merge=read.csv("output/twinstrand_maf_merge.csv",header = T,stringsAsFactors = F)
# twinstrand_maf_merge=read.csv("../output/twinstrand_maf_merge.csv",header = T,stringsAsFactors = F)
ic50data_all=read.csv("data/ic50data/IC50HeatMap.csv",header = T,stringsAsFactors = F)
# ic50data_all=read.csv("../data/IC50HeatMap.csv",header = T,stringsAsFactors = F)
net_gr_wodrug=0.055
conc_for_predictions=0.8

ic50data_all=ic50data_all%>%filter(species%in%c("WT","V299L_H","E355A","D276G_maxi","H396R","F317L","F359I","E459K","G250E","F359C","F359V","M351T","L248V","E355G_maxi","Q252H_maxi","Y253F","F486S_maxi","H396P_maxi","E255K","Y253H","T315I","E255V"))


ic50data_all=ic50data_all%>%
  mutate(species=case_when(species=="F486S_maxi"~"F486S",
                          species=="H396P_maxi"~"H396P",
                          species=="Q252H_maxi"~"Q252H",
                          species=="E355G_maxi"~"E355G",
                          species=="D276G_maxi"~"D276G",
                          species=="V299L_H" ~ "V299L",
                          TRUE ~as.character(species)))

#Looking at how many replicates are there for each mutant in the data
nreplicates=ic50data_all%>%group_by(species)%>%arrange(species)%>%summarize(replicates=n())

#basically adding an identifier for replicate number.
ic50data_all=ic50data_all%>%arrange(species)%>%mutate(replicate=c(1:6,1:6,1:8,1:6,1:6,1:6,1:6,1:6,1:6,1:6,1:6,1:8,1:6,1:6,1:6,1:6,1:6,1:8,1:20,1:6,1:6))

ic50data_all_melt=melt(ic50data_all,id.vars = c("species","replicate"),measure.vars =c("X10","X5","X2.5","X1.25","X0.625","X0.3125","X0.15625","X0.078125","X0.0390625","X0.01953125") ,variable.name = "concentration",value.name = "y")
ic50data_all_melt$concentration=as.character(ic50data_all_melt$concentration)

ic50data_all_melt=ic50data_all_melt%>%rowwise()%>%mutate(concentration=as.numeric(strsplit(concentration,"X")[[1]][2]))

# a=ic50data_all_melt%>%filter(species=="T315I",replicate==1)
# ic50.ll4=drm(y~concentration,data=a,fct=LL.3(fixed=c(NA,1,NA)))
# b=coef(ic50.ll4)[1]
#   c=0
#   d=1
#   e=coef(ic50.ll4)[2]


########################Getting Confidence Intervals########################
###Calculating 4 parameter fits for all datapoints for a given mutant
#Changing 'species' column to 'mutant' column
ic50data_all_melt=ic50data_all_melt%>%mutate(mutant=species)%>%dplyr::select(!species)

ic50data_long=ic50data_all_melt%>%mutate(conc=concentration)
conc.list=c(10,5,2.5,1.25,.625,.3125,.15625,.078125,.0390625,.01953125)
ic50.model.pred=data.frame(matrix(NA,nrow=length(conc.list)*length(ic50data_all$species),ncol=0))

for(species_curr in sort(unique(ic50data_long$mutant))){
  # species_curr="T315I"
  ic50data_species_specific=ic50data_long%>%filter(mutant==species_curr)
  ic50.model.pred.species.specific=data.frame(matrix(NA,nrow=length(conc.list)*length(unique(ic50data_long$mutant)),ncol=0))
  for(rep_curr in sort(unique(ic50data_species_specific$replicate))){
  # rep_curr="1"
  ic50data_species_rep_specific=ic50data_species_specific%>%filter(mutant==species_curr,replicate==rep_curr)
  #Next: Appproximating Response from dose (inverse of the prediction)
  ic50.ll4=drm(y~conc,data=ic50data_species_rep_specific,fct=LL.3(fixed=c(NA,1,NA)))
  #Extracting coefficients
  b=coef(ic50.ll4)[1]
  c=0
  d=1
  e=coef(ic50.ll4)[2]
  # rm(ic50.model.pred.species.rep.specific)
  ic50.model.pred.species.rep.specific=data.frame(matrix(NA,nrow=length(conc.list),ncol=0))
  i=1
  ic50.model.pred.species.rep.specific$mutant=species_curr
  ic50.model.pred.species.rep.specific$replicate=rep_curr
  #For loop for the unique concentrations
  for(conc.curr in conc.list){
    ic50.model.pred.species.rep.specific$conc[i]=conc.curr
    ic50.model.pred.species.rep.specific$y_model[i]=c+((d-c)/(1+exp(b*(log(conc.curr)-log(e)))))
    i=i+1
  }
  ic50.model.pred.species.specific=rbind(ic50.model.pred.species.specific,ic50.model.pred.species.rep.specific)
}
  ic50.model.pred=rbind(ic50.model.pred,ic50.model.pred.species.specific)
}

###Plotting modeled and raw dose responses for T315I and Y253H 
plotly=ggplot(ic50.model.pred,aes(x=conc,y=y_model,color=factor(replicate)))+geom_point()+geom_line()+scale_x_continuous(trans="log10")+facet_wrap(~mutant)
ggplotly(plotly)

plotly=ggplot(ic50.model.pred%>%filter(mutant=="T315I"),aes(x=conc,y=y_model,color=factor(replicate)))+geom_point()+geom_line()+scale_x_continuous(trans="log10")
ggplotly(plotly)

plotly=ggplot(ic50data_long%>%filter(mutant=="T315I"),aes(x=conc,y=y,color=factor(replicate)))+geom_point()+geom_line()+scale_x_continuous(trans="log10")
ggplotly(plotly)

plotly=ggplot(ic50.model.pred%>%filter(mutant=="Y253H"),aes(x=conc,y=y_model,color=factor(replicate)))+geom_point()+geom_line()+scale_x_continuous(trans="log10")
ggplotly(plotly)

plotly=ggplot(ic50data_long%>%filter(mutant=="Y253H"),aes(x=conc,y=y,color=factor(replicate)))+geom_point()+geom_line()+scale_x_continuous(trans="log10")
ggplotly(plotly)

ic50data_all2=ic50.model.pred%>%filter(conc==1.25)%>%dplyr::select(species=mutant,y_model)
    #will just start of with one dose at 1.25uM
    # ic50data_all2=data.frame(cbind(ic50data_all$species,ic50data_all$X1.25))
    # ic50data_all2=data.frame(cbind(ic50data_all$species,ic50data_all$X0.625))
    colnames(ic50data_all2)=c("species","dose_response")
    ic50data_all2$dose_response=as.numeric(as.character(ic50data_all2$dose_response))


    #Calculating confidence limit and standard deviations
    ic50data_all_sum=ic50data_all2%>%group_by(species)%>%summarise(dr_mean=mean(dose_response),dr_ci_ul=dr_mean+1.96*sd(dose_response)*sqrt(n()),dr_ci_ll=dr_mean-1.96*sd(dose_response)*sqrt(n()),dr_sd_ul=dr_mean+sd(dose_response),dr_sd_ll=dr_mean-sd(dose_response))
  

#Since some mutants have 0% alive at the lower limit of the confidence interval, I'm converting those to 0
ic50data_all_sum=ic50data_all_sum%>%mutate(dr_ci_ll=case_when(dr_ci_ll<=0~0,
                                                              TRUE~dr_ci_ll),
                                           dr_sd_ll=case_when(dr_sd_ll<=0~0,
                                                              TRUE~dr_sd_ll))

#Dose response here is essentially y. aka %alive
#Converting y to drug effect on growth rate aka alpha value
ic50data_all_sum=ic50data_all_sum%>%mutate(netgr_pred_mean=net_gr_wodrug-(-log(dr_mean)/72),netgr_pred_ci_ul=net_gr_wodrug-(-log(dr_ci_ul)/72),netgr_pred_ci_ll=net_gr_wodrug-(-log(dr_ci_ll)/72),netgr_pred_sd_ul=net_gr_wodrug-(-log(dr_sd_ul)/72),netgr_pred_sd_ll=net_gr_wodrug-(-log(dr_sd_ll)/72))


# write.csv(ic50data_all_sum,"data/ic50data_all_confidence_intervals_individual_logistic_fits.csv")

```
